---
title: "iSNV_COI_V3"
author: "Alex Di Genova"
date: "2026-05-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## iSNV 

iSNV filters from PASS variants called with bcftools

```{r isnv}
library(tidyverse)
library(broom)
library(patchwork)
library(ggbeeswarm)
library(ggpmisc)
library(ggpubr)
library(glue)

# -----------------------------
# Parameters
# -----------------------------
pars <- list(
  af_min_primary = 0.05,
  af_min_sens    = 0.03,
  af_max         = 0.95,
  alt_qual_min   = 30,
  ct_max         = 25,

  # strand support
  alt_rv_min     = 5,   # at least 5 reverse ALT reads
  alt_fw_min     = 5,   # at least 5 forward ALT reads

  # coarse genome-end mask if no primer BED is available
  end_mask_bp    = 100,

  # recurrent hotspot definition
  recurrent_prop_min = 0.10,   # site seen in >=10% of samples
  recurrent_af_max   = 0.10    # but median AF stays low -> suspicious
)

# -----------------------------
# Helper functions
# -----------------------------
calc_strand_bias_p <- function(ref_dp, ref_rv, alt_dp, alt_rv) {
  ref_fw <- ref_dp - ref_rv
  alt_fw <- alt_dp - alt_rv

  # avoid negative weird cases
  if (any(c(ref_fw, ref_rv, alt_fw, alt_rv) < 0)) return(NA_real_)

  m <- matrix(c(ref_fw, ref_rv, alt_fw, alt_rv), nrow = 2, byrow = TRUE)
  fisher.test(m, alternative = "greater")$p.value
}

read_isnv_file <- function(f) {
  read.table(f, header = TRUE, sep = "\t", stringsAsFactors = FALSE) %>%
    mutate(
      Filename = basename(f),
      barcode  = as.integer(stringr::str_extract(basename(f), "(?<=-)\\d+"))
    )
}

# -----------------------------
# Load raw iSNV candidates
# -----------------------------
files_isnv <- list.files("iSNVs", pattern = "_variants.tsv$", full.names = TRUE)

raw_isnv_cache_file <- "analysis/raw_isnv_cache.rds"
raw_isnv_cache_meta <- tibble(
  file = normalizePath(files_isnv),
  size = file.info(files_isnv)$size,
  mtime = file.info(files_isnv)$mtime
)

if (file.exists(raw_isnv_cache_file)) {
  raw_isnv_cache <- readRDS(raw_isnv_cache_file)
} else {
  raw_isnv_cache <- NULL
}

if (
  !is.null(raw_isnv_cache) &&
  identical(raw_isnv_cache$meta, raw_isnv_cache_meta)
) {
  raw_isnv <- raw_isnv_cache$data
} else {
  raw_isnv <- purrr::map_dfr(files_isnv, read_isnv_file) %>%
    filter(PASS=TRUE) %>%
    mutate(
      ALT_FW = ALT_DP - ALT_RV,
      REF_FW = REF_DP - REF_RV,
      strand_p = purrr::pmap_dbl(
        list(REF_DP, REF_RV, ALT_DP, ALT_RV),
        calc_strand_bias_p
      ),
      both_strands_ok = ALT_RV >= pars$alt_rv_min & ALT_FW >= pars$alt_fw_min,
      basic_ok =
        PASS == TRUE &
        ALT_QUAL >= pars$alt_qual_min &
        ALT_FREQ < pars$af_max,
      genome_end = POS <= pars$end_mask_bp | POS >= (29903 - pars$end_mask_bp)
    )

  saveRDS(
    list(meta = raw_isnv_cache_meta, data = raw_isnv),
    raw_isnv_cache_file
  )
}

# -----------------------------
# Load consensus SNVs
# -----------------------------
files_snv <- list.files("SNVs", pattern = "\\.variants.tsv$", full.names = TRUE)

cons_snv <- purrr::map_dfr(files_snv, \(f) {
  read.table(f, header = TRUE, sep = "\t", stringsAsFactors = FALSE) %>%
    mutate(
      Filename = basename(f),
      barcode  = as.integer(stringr::str_extract(basename(f), "(?<=-)\\d+"))
    )
}) %>%
  transmute(barcode, POS, ALT, cons_key = paste(barcode, POS, ALT, sep = "_"))


#demografic data
df=read.table("demographics/all_samples.txt", h=T)
dates_formatted <- as.Date(df$Date, format="%m/%d/%Y")
years <- format(dates_formatted, "%Y")
df$year=years
df=df %>% mutate(Ct=as.integer(Ct))
df$all="all"

# -----------------------------
# loadcoi: Pangolin selected variant plus COI marker
# -----------------------------
all_results <- read.table("analysis/all_results.csv", h = TRUE, sep = ",") %>%
  mutate(
    barcode = as.integer(stringr::str_extract(sampleId, "(?<=_)\\d+")),
    pangolin_variant = case_when(
      stringr::str_detect(variant, "Omicron") ~ "Omicron",
      TRUE ~ variant
    )
  ) %>%
  transmute(
    barcode,
    pangolin_variant,
    mut_ivar,
    pangolin_location = location,
    pangolin_cluster = cluster
  )

coi_summary_file <- "linages_mutations/coinfection_calls_lineage_v3/cohort_coinfection_summary.tsv"

sample_coi <- read.table(coi_summary_file, h = TRUE, sep = "\t") %>%
  mutate(
    barcode = as.integer(stringr::str_extract(sample, "(?<=-)\\d+")),
    co_infection = co_infection_flag %in% c(TRUE, "TRUE", "True", "true", 1, "1"),
    major_label = na_if(stringr::str_trim(major_label), ""),
    minor_label = na_if(stringr::str_trim(minor_label), ""),
    coi_major_lineage = major_label,
    coi_minor_lineage = if_else(co_infection, minor_label, NA_character_),
    coi_lineage_call = if_else(
      co_infection,
      paste(major_label, minor_label, sep = " + "),
      major_label
    )
  ) %>%
  transmute(
    barcode,
    coi_major_lineage,
    coi_major_frac_supported_sites = major_frac_supported_sites,
    coi_major_status = major_status,
    coi_major_n_supported_sites = major_n_supported_sites,
    coi_minor_lineage,
    coi_minor_n_supported_sites = minor_n_supported_sites,
    coi_minor_median_af = minor_median_af,
    coi_confidence_tier = confidence_tier,
    co_infection,
    coi_lineage_call
  )

df <- df %>%
  left_join(all_results, by = "barcode") %>%
  left_join(sample_coi, by = "barcode") %>%
  mutate(
    pangolin_variant = replace_na(pangolin_variant, "Unknown"),
    coi_lineage_call = replace_na(coi_lineage_call, "Unknown"),
    coi_confidence_tier = replace_na(coi_confidence_tier, "none"),
    co_infection = replace_na(co_infection, FALSE),
    coi_major_status = replace_na(coi_major_status, "unassigned"),
    lineage_group = if_else(co_infection, "Co-Infection", pangolin_variant)
  )



sample_qc <- df %>%
  select(barcode, Ct, pangolin_variant, mut_ivar, pangolin_location, pangolin_cluster,
         coi_major_lineage, coi_major_frac_supported_sites, coi_major_status,
         coi_major_n_supported_sites, coi_minor_lineage, coi_minor_n_supported_sites,
         coi_minor_median_af, coi_confidence_tier, co_infection, coi_lineage_call,
         lineage_group) %>%
  mutate(
    sample_ok = Ct <= pars$ct_max
  )

# -----------------------------
# Annotate candidate calls
# -----------------------------
raw_isnv <- raw_isnv %>%
  mutate(cons_key = paste(barcode, POS, ALT, sep = "_")) %>%
  left_join(sample_qc, by = "barcode") %>%
  mutate(
    is_consensus_snv = cons_key %in% cons_snv$cons_key,
    site_key = paste(REGION, POS, REF, ALT, sep = "_")
  )

# -----------------------------
# Recurrent suspicious hotspots
# Seen in many samples but with low median AF
# -----------------------------
n_samples <- raw_isnv %>% distinct(barcode) %>% nrow()

hotspots <- raw_isnv %>%
  filter(
    basic_ok,
    ALT_FREQ >= pars$af_min_sens,
    !is.na(sample_ok), sample_ok,
    !is_consensus_snv
  ) %>%
  group_by(site_key, REGION, POS, REF, ALT) %>%
  summarise(
    n_samples_site = n_distinct(barcode),
    prop_samples   = n_samples_site / n_samples,
    median_af      = median(ALT_FREQ, na.rm = TRUE),
    mean_alt_rv    = mean(ALT_RV, na.rm = TRUE),
    mean_alt_fw    = mean(ALT_FW, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    suspicious_hotspot =
      prop_samples >= pars$recurrent_prop_min &
      median_af <= pars$recurrent_af_max
  )

raw_isnv <- raw_isnv %>%
  left_join(
    hotspots %>% select(site_key, suspicious_hotspot),
    by = "site_key"
  ) %>%
  mutate(suspicious_hotspot = replace_na(suspicious_hotspot, FALSE))

# -----------------------------
# Final primary and sensitivity filters
# -----------------------------
isnv_primary <- raw_isnv %>%
  filter(
    basic_ok,
    ALT_FREQ >= pars$af_min_primary,
    !is.na(sample_ok), sample_ok,
    !is_consensus_snv,
    both_strands_ok,
    !genome_end,
    !suspicious_hotspot
  )

isnv_sensitivity <- raw_isnv %>%
  filter(
    basic_ok,
    ALT_FREQ >= pars$af_min_sens,
    !is.na(sample_ok), sample_ok,
    !is_consensus_snv,
    both_strands_ok,
    !genome_end,
    !suspicious_hotspot
  )

```

## iSNV plots

summary of iSNV passing variants per filter

```{r pressure, echo=FALSE}

library(tidyverse)
library(scales)

qc_counts <- tibble(
  step = c(
    "Raw",
    "AF >= 3%",
    "Ct <= 28",
    "Not consensus SNV",
    "Both strands",
    "Not genome end",
    "Not recurrent hotspot",
    "Primary AF >= 5%"
  ),
  n = c(
    nrow(raw_isnv),
    sum(raw_isnv$basic_ok & raw_isnv$ALT_FREQ >= pars$af_min_sens, na.rm = TRUE),
    sum(raw_isnv$basic_ok & raw_isnv$ALT_FREQ >= pars$af_min_sens & raw_isnv$sample_ok, na.rm = TRUE),
    sum(raw_isnv$basic_ok & raw_isnv$ALT_FREQ >= pars$af_min_sens & raw_isnv$sample_ok & !raw_isnv$is_consensus_snv, na.rm = TRUE),
    sum(raw_isnv$basic_ok & raw_isnv$ALT_FREQ >= pars$af_min_sens & raw_isnv$sample_ok & !raw_isnv$is_consensus_snv & raw_isnv$both_strands_ok, na.rm = TRUE),
    sum(raw_isnv$basic_ok & raw_isnv$ALT_FREQ >= pars$af_min_sens & raw_isnv$sample_ok & !raw_isnv$is_consensus_snv & raw_isnv$both_strands_ok & !raw_isnv$genome_end, na.rm = TRUE),
    sum(raw_isnv$basic_ok & raw_isnv$ALT_FREQ >= pars$af_min_sens & raw_isnv$sample_ok & !raw_isnv$is_consensus_snv & raw_isnv$both_strands_ok & !raw_isnv$genome_end & !raw_isnv$suspicious_hotspot, na.rm = TRUE),
    nrow(isnv_primary)
  )
) %>%
  mutate(
    step = factor(step, levels = rev(step)),
    retained_pct = 100 * n / n[1],
    label = paste0(comma(n), " (", sprintf("%.1f", retained_pct), "%)"),
    group = case_when(
      step == "Raw" ~ "Raw",
      step == "Primary AF >= 5%" ~ "Final",
      TRUE ~ "Intermediate"
    )
  )

a=ggplot(qc_counts, aes(x = step, y = n, fill = group)) +
  geom_col(width = 0.75) +
  geom_text(
    aes(label = label),
    hjust = -0.08,
    size = 5
  ) +
  coord_flip(clip = "off") +
  scale_y_log10(
    labels = comma,
    expand = expansion(mult = c(0, 0.28))
  ) +
  scale_fill_manual(
    values = c(
      "Raw" = "#4D4D4D",
      "Intermediate" = "#4C78A8",
      "Final" = "#2E8B57"
    )
  ) +
  labs(
   # title = "iSNV filtering cascade",
    x = NULL,
    y = "Number of candidate iSNVs (log10 scale)"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    legend.position = "none",
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 14),
    axis.text.x = element_text(size = 13),
    plot.title = element_text(size = 24, face = "bold"),
    axis.title.x = element_text(size = 16),
    plot.margin = margin(20, 120, 20, 20)
  )

a
pdf("Filtering_iSNV_steps.pdf",7,4)
a
dev.off()
```


## Primary iSNV variable diagnostics

These plots inspect the retained `isnv_primary` calls directly. They are useful for checking
whether the final calls have plausible allele frequencies, read depths, strand support,
quality values, genomic distribution, and mutation spectra.

qc V2

```{r qcv2}
library(tidyverse)
library(patchwork)
library(scales)
library(viridis)

# ============================================================
# Main iSNV QC figure
# ============================================================

# -----------------------------
# Sample-level summary
# -----------------------------
isnv_counts <- isnv_primary %>%
  group_by(barcode) %>%
  summarise(
    n_isnv = n(),
    mean_alt_freq = mean(ALT_FREQ, na.rm = TRUE),
    median_alt_freq = median(ALT_FREQ, na.rm = TRUE),
    mean_total_dp = mean(TOTAL_DP, na.rm = TRUE),
    median_total_dp = median(TOTAL_DP, na.rm = TRUE),
    .groups = "drop"
  )

isnv_sample <- df %>%
  mutate(
    barcode = as.integer(barcode),
    Date = as.Date(Date, format = "%m/%d/%Y"),
    Sex = factor(Sex),
    pangolin_variant = factor(pangolin_variant),
    co_infection = factor(
      co_infection,
      levels = c(FALSE, TRUE),
      labels = c("No", "Yes")
    ),
    lineage_group = factor(lineage_group),
    sample_ok = Ct <= pars$ct_max
  ) %>%
  left_join(isnv_counts, by = "barcode") %>%
  mutate(
    n_isnv = replace_na(n_isnv, 0L),
    any_isnv = n_isnv > 0
  )

isnv_sample_qc <- isnv_sample %>%
  filter(sample_ok)

# -----------------------------
# Call-level summary
# -----------------------------
isnv_primary_plot <- isnv_primary %>%
  mutate(
    mutation = paste0(REF, ">", ALT),
    strand_balance = ALT_FW / ALT_DP
  )

mutation_order <- c(
  "A>C", "A>G", "A>T",
  "C>A", "C>G", "C>T",
  "G>A", "G>C", "G>T",
  "T>A", "T>C", "T>G"
)

isnv_primary_plot <- isnv_primary_plot %>%
  mutate(
    mutation = factor(mutation, levels = mutation_order)
  )

# -----------------------------
# Summary stats
# -----------------------------
n_samples <- nrow(isnv_sample_qc)
n_with_isnv <- sum(isnv_sample_qc$n_isnv > 0)
med_isnv <- median(isnv_sample_qc$n_isnv)
iqr_isnv <- IQR(isnv_sample_qc$n_isnv)
total_primary <- nrow(isnv_primary)

# -----------------------------
# Colors and theme
# -----------------------------
main_blue <- "#4C78A8"
threshold_red <- "#C44E52"
main_grey <- "grey35"

mutation_cols <- c(
  "A>C" = "#8FB9A8",
  "A>G" = "#D8A25E",
  "A>T" = "#B9C36A",
  "C>A" = "#7EA3B0",
  "C>G" = "#A79BBE",
  "C>T" = "#2A9D8F",
  "G>A" = "#6FAE9B",
  "G>C" = "#8DB7C7",
  "G>T" = "#4C78A8",
  "T>A" = "#9C7AA5",
  "T>C" = "#C97AA4",
  "T>G" = "#E07A5F"
)

theme_qc <- theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 12.5),
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = 11.5),
    axis.text = element_text(size = 10.5),
    legend.title = element_text(size = 10.5),
    legend.text = element_text(size = 9.5)
  )

# -----------------------------
# A. Burden per sample
# -----------------------------
p1 <- ggplot(isnv_sample_qc, aes(x = n_isnv)) +
  geom_histogram(
    binwidth = 1,
    boundary = -0.5,
    fill = main_grey,
    color = "white",
    linewidth = 0.2
  ) +
  geom_vline(
    xintercept = med_isnv,
    linetype = "dashed",
    color = threshold_red,
    linewidth = 0.8
  ) +
  labs(
    title = "Burden per sample",
    x = "Primary iSNVs per sample",
    y = "Samples"
  ) +
  theme_qc

# -----------------------------
# B. Burden versus Ct
# -----------------------------
p2 <- ggplot(isnv_sample_qc, aes(x = Ct, y = n_isnv)) +
  geom_point(
    color = main_blue,
    alpha = 0.75,
    size = 2.2
  ) +
  geom_smooth(
    method = "glm",
    method.args = list(family = "poisson"),
    se = TRUE,
    color = "black",
    linewidth = 0.8
  ) +
  labs(
    title = "Burden vs Ct",
    x = "Ct",
    y = "Primary iSNVs per sample"
  ) +
  theme_qc

# -----------------------------
# C. ALT frequency distribution
# -----------------------------
p3 <- ggplot(isnv_primary_plot, aes(x = ALT_FREQ)) +
  geom_histogram(
    binwidth = 0.025,
    fill = main_blue,
    color = "white",
    linewidth = 0.2
  ) +
  geom_vline(
    xintercept = pars$af_min_primary,
    linetype = "dashed",
    color = threshold_red,
    linewidth = 0.8
  ) +
  labs(
    title = "ALT frequency",
    x = "ALT frequency",
    y = "Primary iSNV count"
  ) +
  theme_qc

# -----------------------------
# D. Depth support
# -----------------------------
p4 <- ggplot(
  isnv_primary_plot,
  aes(x = TOTAL_DP, y = ALT_DP, color = ALT_FREQ)
) +
  geom_point(
    alpha = 0.55,
    size = 1.5
  ) +
  geom_abline(
    intercept = 0,
    slope = pars$af_min_primary,
    linetype = "dashed",
    color = threshold_red,
    linewidth = 0.8
  ) +
  scale_x_log10(labels = comma) +
  scale_y_log10(labels = comma) +
  scale_color_viridis_c(
    option = "C",
    name = "AF",
    end = 0.95
  ) +
  labs(
    title = "Depth support",
    x = "Total depth",
    y = "ALT depth"
  ) +
  theme_qc

# -----------------------------
# E. Genome-wide distribution
# -----------------------------
p5 <- ggplot(isnv_primary_plot, aes(x = POS)) +
  geom_histogram(
    binwidth = 500,
    fill = main_blue,
    color = "white",
    linewidth = 0.2
  ) +
  annotate(
    "rect",
    xmin = 1,
    xmax = pars$end_mask_bp,
    ymin = -Inf,
    ymax = Inf,
    fill = threshold_red,
    alpha = 0.12
  ) +
  annotate(
    "rect",
    xmin = 29903 - pars$end_mask_bp,
    xmax = 29903,
    ymin = -Inf,
    ymax = Inf,
    fill = threshold_red,
    alpha = 0.12
  ) +
  labs(
    title = "Genome distribution",
    x = "Genome position",
    y = "Primary iSNV count"
  ) +
  theme_qc

# -----------------------------
# F. Mutation spectrum
# -----------------------------
p6 <- ggplot(isnv_primary_plot, aes(x = mutation, fill = mutation)) +
  geom_bar(
    color = "white",
    linewidth = 0.2
  ) +
  scale_fill_manual(values = mutation_cols, drop = FALSE) +
  labs(
    title = "Mutation spectrum",
    x = "Mutation type",
    y = "Primary iSNV count"
  ) +
  theme_qc +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# -----------------------------
# Combine figure
# -----------------------------
qc_results_fig <-
  (p1 | p2) /
  (p3 | p4) /
  (p5 | p6) +
  plot_annotation(
   # title = "Primary iSNV quality control",
    tag_levels = "A",
    theme = theme(
      plot.title = element_text(size = 16, face = "bold")
    )
  )

qc_results_fig

# -----------------------------
# Save figure
# -----------------------------
# ggsave(
#   "figures/isnv_results_qc_main.png",
#   qc_results_fig,
#   width = 12,
#   height = 11,
#   dpi = 300
# )
# 
ggsave(
  "isnv_results_qc_main.pdf",
  qc_results_fig,
  width = 12,
  height = 11
)
```

## Comparison iSNV vs SNV

This reproduces the SNV/iSNV comparison from `MFig_Covid.Rmd` using the cleaned
`isnv_primary` calls and keeping zero-count samples in both burdens.

```{r cmpsnvs_clean, fig.width=12, fig.height=7}
MutSNV <- cons_snv %>%
  count(barcode, name = "n_snv") %>%
  right_join(df, by = "barcode") %>%
  mutate(
    n_snv = replace_na(n_snv, 0L),
    Date = as.Date(Date, format = "%m/%d/%Y")
  ) %>%
  filter(Ct <= pars$ct_max)

iSNV_date_data <- isnv_sample_qc %>%
  mutate(Date = as.Date(Date, format = "%m/%d/%Y"))

add_coi_medium_plot_group <- function(dat) {
  base_dat <- dat %>%
    mutate(lineage_plot_group = as.character(lineage_group))

  medium_coi_dat <- base_dat %>%
    filter(lineage_group == "Co-Infection", coi_confidence_tier == "medium") %>%
    mutate(lineage_plot_group = "Co-Infection medium")

  plot_dat <- bind_rows(base_dat, medium_coi_dat)
  preferred_levels <- c("Other", "Lambda", "Omicron", "Co-Infection", "Co-Infection medium")
  extra_levels <- setdiff(sort(unique(plot_dat$lineage_plot_group)), preferred_levels)

  plot_dat %>%
    mutate(
      lineage_plot_group = factor(
        lineage_plot_group,
        levels = c(preferred_levels, extra_levels)
      )
    )
}

MutSNV_plot <- add_coi_medium_plot_group(MutSNV)
iSNV_date_plot <- add_coi_medium_plot_group(iSNV_date_data)

p_snv_sex <- MutSNV_plot %>%
  ggplot(aes(x = Sex, y = n_snv, color = pangolin_location)) +
  geom_quasirandom(width = 0.18, alpha = 0.85) +
     scale_color_manual(values = c("Santiago" = "blue", "Rancagua" = "red")) + 
  facet_wrap(~ year) +
  labs(x = "Sexo", y = "SNV", color = "Variant group", title = "SNVs per year") +
  theme_light() +
  theme(legend.position = "none")

p_isnv_sex <- iSNV_date_plot %>%
  ggplot(aes(x = Sex, y = n_isnv, color = pangolin_location)) +
  geom_quasirandom(width = 0.18, alpha = 0.85) +
   scale_color_manual(values = c("Santiago" = "blue", "Rancagua" = "red")) + 
  facet_wrap(~ year) +
  labs(x = "Sexo", y = "iSNV", color = "Variant group", title = "iSNVs per year") +
  theme_light() +
  theme(legend.position = "none")

p_snv_date <- MutSNV_plot %>%
  ggplot(aes(x = Date, y = n_snv)) +
  geom_point(aes(color = pangolin_location), size = 2, alpha = 0.9) +
  geom_smooth(data = MutSNV, method = "lm", se = TRUE) +
   scale_color_manual(values = c("Santiago" = "blue", "Rancagua" = "red")) + 
  stat_poly_eq(
    data = MutSNV,
    aes(label = paste(after_stat(eq.label), after_stat(rr.label), sep = "~~~")),
    formula = y ~ x,
    parse = TRUE
  ) +
  scale_x_date(date_breaks = "4 months", date_labels = "%b/%Y") +
  labs(title = "SNVs vs Date", x = "Date", y = "SNV") +
  theme_light() +
  theme(legend.position = "none")

p_isnv_date <- iSNV_date_plot %>% 
  ggplot(aes(x = Date, y = n_isnv)) +
  geom_point(aes(color = Region), size = 2, alpha = 0.9) +
  geom_smooth(data = iSNV_date_data, method = "lm", se = TRUE) +
   scale_color_manual(values = c("Metropolitana" = "blue", "O'Higgins" = "red")) + 
  stat_poly_eq(
    data = iSNV_date_data,
    aes(label = paste(after_stat(eq.label), after_stat(rr.label), sep = "~~~")),
    formula = y ~ x,
    parse = TRUE
  ) +
  scale_x_date(date_breaks = "4 months", date_labels = "%b/%Y") +
  labs(title = "iSNVs vs Date", x = "Date", y = "iSNV") +
  theme_light() +
  theme(legend.position = "bottom")

(p_snv_sex / p_isnv_sex) | (p_snv_date / p_isnv_date)

pdf("iSNV_SNV_date.pdf",8,6)
(p_snv_sex / p_isnv_sex) | (p_snv_date / p_isnv_date)
dev.off()


```

```{r isnv_snv_date_lm}
snv_date_lm <- lm(n_snv ~ Date, data = MutSNV)
isnv_date_lm <- lm(n_isnv ~ Date, data = iSNV_date_data)

summary(snv_date_lm)
summary(isnv_date_lm)

glimpse(isnv_date_lm)
```

## Linear Regression Model

This linear model using iSNV burden and Pangolin/COI annotations from this analysis.

```{r isnv_linear_model}
coverage_file <- if (file.exists("resume_samtools_coverage.txt")) {
  "resume_samtools_coverage.txt"
} else {
  "analysis/resume_samtools_coverage.txt"
}

coverage_qc <- read.table(coverage_file, h = TRUE) %>%
  mutate(
    barcode = as.integer(stringr::str_extract(sname, "(?<=-)\\d+")),
    logc = log10(meandepth)
  ) %>%
  select(barcode, meandepth, coverage, logc)

variant_min_n <- 5

valid_lineage_groups <- isnv_sample_qc %>%
  count(lineage_group, name = "n_samples_group") %>%
  filter(n_samples_group >= variant_min_n) %>%
  pull(lineage_group)

isnv_lm_data <- isnv_sample_qc %>%
  filter(lineage_group %in% valid_lineage_groups) %>%
  left_join(coverage_qc, by = "barcode") %>%
  mutate(
    age2 = case_when(
      Age < 30 ~ "young",
      Age >= 30 & Age < 50 ~ "adult",
      TRUE ~ "older"
    ),
    age2 = relevel(factor(age2), ref = "adult"),
    Sex = factor(Sex),
    lineage_group = relevel(droplevels(factor(lineage_group)), ref = "Other"),
    n = n_isnv
  ) %>%
  drop_na(n, age2, Sex, mut_ivar, lineage_group, Ct, logc)

isnv_lm_data %>%
  count(lineage_group, name = "n_samples_group") %>%
  arrange(desc(n_samples_group))

iSNVlm <- lm(n ~ Age + Sex + mut_ivar + lineage_group + Ct + logc, data = isnv_lm_data)
summary(iSNVlm)
```

```{r negative_binomial}

library(MASS)
library(broom)

m_nb_adj <- glm.nb(
  n_isnv ~ Ct + logc + pangolin_variant + Age + Sex + co_infection,
  data = isnv_lm_data
)

tidy(m_nb_adj, exponentiate = TRUE, conf.int = TRUE)



```
boostrap model 
```{r bostrap}

library(MASS)
library(boot)
library(dplyr)

# Fit the original model first
m_nb_adj <- glm.nb(
  n_isnv ~ Ct + logc + pangolin_variant + Age + Sex + co_infection,
  data = isnv_lm_data
)

# Expected coefficient names
coef_names <- names(coef(m_nb_adj))

# Safe bootstrap function
boot_fun <- function(data, idx) {
  d <- data[idx, ]

  fit <- tryCatch(
    glm.nb(
      n_isnv ~ Ct + logc + pangolin_variant + Age + Sex + co_infection,
      data = d
    ),
    error = function(e) NULL
  )

  out <- rep(NA_real_, length(coef_names))
  names(out) <- coef_names

  if (!is.null(fit)) {
    b <- coef(fit)
    out[names(b)] <- b
  }

  out
}

set.seed(123)

boot_nb <- boot(
  data = isnv_lm_data,
  statistic = boot_fun,
  R = 1000
)

boot_nb


boot_summary <- lapply(seq_along(coef_names), function(i) {
  vals <- boot_nb$t[, i]
  vals <- vals[is.finite(vals)]

  tibble(
    term = coef_names[i],
    estimate_log = coef(m_nb_adj)[i],
    boot_low_log = quantile(vals, 0.025, na.rm = TRUE),
    boot_high_log = quantile(vals, 0.975, na.rm = TRUE),
    estimate = exp(coef(m_nb_adj)[i]),
    boot_low = exp(quantile(vals, 0.025, na.rm = TRUE)),
    boot_high = exp(quantile(vals, 0.975, na.rm = TRUE)),
    n_success = length(vals)
  )
}) %>%
  bind_rows()

boot_summary

boot_summary %>%
  filter(term %in% c(
    "SexM",
    "co_infectionYes",
    "pangolin_variantOmicron",
    "Ct",
    "logc"
  ))
```

bayesian model
```{r bayes}
library(brms)

isnv_lm_data2 <- isnv_lm_data %>%
  mutate(
    z_Ct = as.numeric(scale(Ct)),
    z_log_depth = as.numeric(scale(log10(meandepth))),
    z_Age = as.numeric(scale(Age))
  )

m_bayes_scaled <- brm(
  n_isnv ~ z_Ct + z_log_depth + pangolin_variant + z_Age + Sex + co_infection,
  family = negbinomial(),
  data = isnv_lm_data2,
  chains = 4,
  cores = 4,
  iter = 6000,
  warmup = 3000,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 123
)

summary(m_bayes_scaled)
posterior_summary(m_bayes_scaled)

```

pairs comparison
```{r krustal}

# Variants/groups with enough samples
kw_data <- isnv_lm_data %>%
  filter(lineage_group %in% c("Lambda", "Omicron", "Other", "Co-Infection"))

kruskal.test(n_isnv ~ lineage_group, data = kw_data)

pairwise.wilcox.test(
  kw_data$n_isnv,
  kw_data$lineage_group,
  p.adjust.method = "BH",
  exact = FALSE
)

# -----------------------------
# Sex comparison
# -----------------------------
sex_summary <- isnv_lm_data %>%
  group_by(Sex) %>%
  summarise(
    n = n(),
    median_iSNV = median(n_isnv, na.rm = TRUE),
    IQR_iSNV = IQR(n_isnv, na.rm = TRUE),
    mean_iSNV = mean(n_isnv, na.rm = TRUE),
    .groups = "drop"
  )

sex_summary

# Wilcoxon rank-sum test
wilcox.test(n_isnv ~ Sex, data = isnv_lm_data, exact = FALSE)

```
generate the figure for the ms figure version v2 
```{r fmethods}
library(tidyverse)
library(brms)
library(patchwork)
library(ggpubr)
library(scales)
library(tibble)

# ============================================================
# Figure: Bayesian predictors + unadjusted group comparisons
# ============================================================

# ------------------------------------------------------------
# Panel A: Bayesian model forest plot
# ------------------------------------------------------------

bayes_tab <- as.data.frame(
  fixef(m_bayes_scaled, probs = c(0.025, 0.975))
) %>%
  tibble::rownames_to_column("term") %>%
  as_tibble() %>%
  filter(term != "Intercept") %>%
  mutate(
    estimate = exp(Estimate),
    conf.low = exp(Q2.5),
    conf.high = exp(Q97.5)
  )

forest_dat <- bayes_tab %>%
  mutate(
    term_label = dplyr::case_when(
      term == "z_Ct" ~ "PCR Ct value (1 SD)",
      term == "z_log_depth" ~ "Average genome coverage (log, 1 SD)",
      term == "pangolin_variantGamma" ~ "Variant: Gamma",
      term == "pangolin_variantLambda" ~ "Variant: Lambda",
      term == "pangolin_variantOmicron" ~ "Variant: Omicron",
      term == "pangolin_variantOther" ~ "Variant: Other",
      term == "z_Age" ~ "Age (1 SD)",
      term == "SexM" ~ "Sex: male (vs female)",
      term == "co_infectionYes" ~ "Candidate co-infection",
      TRUE ~ term
    )
  ) %>%
  dplyr::select(term_label, estimate, conf.low, conf.high) %>%
  arrange(estimate) %>%
  mutate(
    term_label = factor(term_label, levels = term_label)
  )

pA <- ggplot(forest_dat, aes(x = estimate, y = term_label)) +
  geom_vline(
    xintercept = 1,
    linetype = "dashed",
    color = "grey45",
    linewidth = 0.7
  ) +
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high),
    height = 0.18,
    linewidth = 0.8,
    color = "black"
  ) +
  geom_point(
    size = 2.9,
    color = "black"
  ) +
  scale_x_log10(
    breaks = c(0.5, 0.75, 1, 1.25, 1.5, 2),
    labels = scales::label_number(accuracy = 0.01)
  ) +
  labs(
    title = "Predictors of iSNV count (IRR, 95% CrI)",
    x = "Incidence rate ratio",
    y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    axis.title.x = element_text(size = 13),
    axis.text.y = element_text(size = 11.5),
    axis.text.x = element_text(size = 11),
    panel.grid.minor = element_blank(),
    plot.margin = margin(t = 8, r = 12, b = 8, l = 8)
  )

# ------------------------------------------------------------
# Panel B: Sex comparison, unadjusted Wilcoxon test
# ------------------------------------------------------------

sex_dat <- isnv_lm_data %>%
  mutate(
    Sex = factor(Sex, levels = c("F", "M"))
  )

ymin_B <- min(sex_dat$n_isnv, na.rm = TRUE)
ymax_B <- max(sex_dat$n_isnv, na.rm = TRUE)

pB <- ggplot(sex_dat, aes(x = Sex, y = n_isnv, color = Sex)) +
  geom_boxplot(
    outlier.shape = NA,
    width = 0.45,
    alpha = 0.25,
    linewidth = 0.8
  ) +
  geom_jitter(
    width = 0.12,
    size = 2.2,
    alpha = 0.8
  ) +
  stat_compare_means(
    method = "wilcox.test",
    label = "p.format",
    label.y = ymax_B + 12,
    size = 5
  ) +
  scale_color_manual(
    values = c(
      "F" = "#4C78A8",
      "M" = "#E0B52B"
    )
  ) +
  scale_y_continuous(
    limits = c(ymin_B - 2, ymax_B + 20),
    expand = expansion(mult = c(0.02, 0.02))
  ) +
  coord_cartesian(clip = "off") +
  labs(
    x = "Sex",
    y = "Number of iSNVs"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 13),
    axis.text = element_text(size = 12),
    panel.grid.minor = element_blank(),
    plot.margin = margin(t = 22, r = 10, b = 8, l = 8)
  )

# ------------------------------------------------------------
# Panel C: Lineage-group comparison
# Kruskal-Wallis + selected pairwise Wilcoxon tests
# ------------------------------------------------------------

kw_data <- isnv_lm_data %>%
  filter(lineage_group %in% c("Other", "Lambda", "Omicron", "Co-Infection")) %>%
  mutate(
    lineage_group = factor(
      lineage_group,
      levels = c("Other", "Lambda", "Omicron", "Co-Infection")
    )
  ) %>%
  droplevels()

ymin_C <- min(kw_data$n_isnv, na.rm = TRUE)
ymax_C <- max(kw_data$n_isnv, na.rm = TRUE)

# Pairwise comparisons selected from your BH-adjusted Wilcoxon results
comparisons <- list(
  c("Other", "Omicron"),
  c("Lambda", "Omicron"),
  c("Other", "Co-Infection")
)

pC <- ggplot(kw_data, aes(x = lineage_group, y = n_isnv, color = lineage_group)) +
  geom_boxplot(
    outlier.shape = NA,
    width = 0.5,
    alpha = 0.22,
    linewidth = 0.8
  ) +
  geom_jitter(
    width = 0.12,
    size = 2.2,
    alpha = 0.85
  ) +
  stat_compare_means(
    method = "kruskal.test",
    label = "p.format",
    label.y = ymax_C + 40,
    size = 4.7
  ) +
  stat_compare_means(
    comparisons = comparisons,
    method = "wilcox.test",
    p.adjust.method = "BH",
    label = "p.format",
    label.y = c(ymax_C + 8, ymax_C + 18, ymax_C + 28),
    size = 4.2
  ) +
  scale_color_manual(
    values = c(
      "Other" = "#3A9D23",
      "Lambda" = "#8E7CC3",
      "Omicron" = "#C67C2E",
      "Co-Infection" = "black"
    )
  ) +
  scale_y_continuous(
    limits = c(ymin_C - 2, ymax_C + 48),
    expand = expansion(mult = c(0.02, 0.02))
  ) +
  coord_cartesian(clip = "off") +
  labs(
    x = "SARS-CoV-2 variant",
    y = "Number of iSNVs"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 13),
    axis.text = element_text(size = 12),
    axis.text.x = element_text(angle = 20, hjust = 1),
    panel.grid.minor = element_blank(),
    plot.margin = margin(t = 28, r = 10, b = 8, l = 8)
  )

# ------------------------------------------------------------
# Combine figure
# ------------------------------------------------------------

final_fig <- pA + (pB / pC) +
  plot_layout(widths = c(1.25, 1)) +
  plot_annotation(
    tag_levels = "A",
    theme = theme(
      plot.tag = element_text(face = "bold", size = 18)
    )
  )

final_fig

# ------------------------------------------------------------
# Save figure
# ------------------------------------------------------------

#ggsave(
#  "figures/isnv_predictors_bayes_unadjusted.png",
#  final_fig,
#  width = 13,
#  height = 7.5,
#  dpi = 300
#)

ggsave(
  "isnv_predictors_bayes_unadjusted.pdf",
  final_fig,
  width = 10,
  height = 7.5
)
```

## DN/DS analysis

we perform primary analysis of dn/ds using the primary data first and new CO_Infection calls
```{r dnds}

library(dndscv)
#sampleID chr      pos ref mut
names <- c(sampleID="barcode", chr="REGION", pos="POS",ref="REF",mut="ALT")
omicron=isnv_primary %>% filter(Ct<=25) %>% filter(lineage_group=="Omicron") %>% dplyr::select(barcode,REGION,POS,REF,ALT) %>% rename(all_of(names))


#we remove iSNVs happening at the same positiion within
other=isnv_primary %>% filter(Ct<=25) %>% filter(lineage_group=="Other") %>% dplyr::select(barcode,REGION,POS,REF,ALT) %>% rename(all_of(names)) #%>%  distinct(chr, pos, ref, mut, .keep_all = TRUE)
Lambda=isnv_primary %>% filter(Ct<=25) %>% filter(lineage_group=="Lambda") %>% dplyr::select(barcode,REGION,POS,REF,ALT) %>% rename(all_of(names)) #%>% distinct(chr, pos, ref, mut, .keep_all = TRUE)
Coi=isnv_primary %>% filter(Ct<=25) %>% filter(lineage_group=="Co-Infection") %>% dplyr::select(barcode,REGION,POS,REF,ALT) %>% rename(all_of(names)) #%>% distinct(chr, pos, ref, mut, .keep_all = TRUE)

# we run the dnds analisys 
dnds_omi = dndscv(omicron, refdb = "analysis/dndscv/db/RefCDS_MN908947.3.peptides.rda", max_muts_per_gene_per_sample = Inf, max_coding_muts_per_sample = Inf)
dnds_oth = dndscv(other, refdb = "analysis/dndscv/db/RefCDS_MN908947.3.peptides.rda", max_muts_per_gene_per_sample = Inf, max_coding_muts_per_sample = Inf)
dnds_lam = dndscv(Lambda, refdb = "analysis/dndscv/db/RefCDS_MN908947.3.peptides.rda", max_muts_per_gene_per_sample = Inf, max_coding_muts_per_sample = Inf)
dnds_coi = dndscv(Coi, refdb = "analysis//dndscv/db/RefCDS_MN908947.3.peptides.rda", max_muts_per_gene_per_sample = Inf, max_coding_muts_per_sample = Inf)

a=dnds_oth$sel_cv %>% dplyr::select(gene_name,wmis_cv) %>% mutate(var="Other")
b=dnds_lam$sel_cv %>% dplyr::select(gene_name,wmis_cv) %>% mutate(var="Lambda")
c=dnds_omi$sel_cv %>% dplyr::select(gene_name,wmis_cv) %>% mutate(var="Omicron")
d=dnds_coi$sel_cv %>% dplyr::select(gene_name,wmis_cv) %>% mutate(var="Co-Infection")

```


plot the results

```{r figdnds2}
# ---- Packages
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(patchwork)


# ---- 3) Aesthetic choices ----
variant_colors <- c(
  "Lambda" = "#8E7CC3",
  "Omicron" = "#B45F06",
  "Other" = "#38761D",
  "Co-Infection" = "#333333"
)

# ---- Global panel with CIs
global_dnds <- dplyr::bind_rows(
  dnds_oth$globaldnds[5,],
  dnds_lam$globaldnds[5,],
  dnds_omi$globaldnds[5,],
  dnds_coi$globaldnds[5,]
) %>%
  mutate(
    name = c("Other","Lambda","Omicron","Co-Infection"),
    name = factor(name, levels = c("Other","Lambda","Omicron","Co-Infection"))
  )

p_global <- ggplot(global_dnds, aes(x = name, y = mle, color = name)) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_pointrange(aes(ymin = cilow, ymax = cihigh), position = position_dodge(width = 0.2)) +
  scale_color_manual(values = variant_colors, guide = "none") + ylim(0,1.2) +
#  coord_flip() +
  labs(
    x = NULL,
    y = expression("Global " * omega ~ "(dN/dS, MLE)"),
    title = "Global dN/dS"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(size = 12),
    axis.title.y = element_text(margin = margin(r = 5)),
    axis.text.x = element_text(size = 10, angle = 20, hjust = 1)
  )


# ---- Combine (narrow global panel + wider gene panel)
#final_dn_ds_fig <- p_global + p_genes + plot_layout(widths = c(1, 3))
#final_dn_ds_fig
#pdf("FigX.pdf",height = 8,width = 16)
#final_dn_ds_fig
#dev.off()


# ---- 1) Combine & harmonize ----
# a, b, c, d must each have: gene_name, wmis_cv, var
df <- dplyr::bind_rows(a, b, c, d) %>%
  mutate(
    var = factor(var, levels = c("Other", "Lambda", "Omicron", "Co-Infection"))
  )

# ---- 2) Global gene order (same across all panels) ----
# Order by max dN/dS across variants (descending)
gene_levels <- df %>%
  group_by(gene_name) %>%
  summarize(max_omega = max(wmis_cv, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(max_omega)) %>%
  pull(gene_name)

df <- df %>%
  mutate(
    gene_name = factor(gene_name, levels = rev(gene_levels)),   # top = highest
    # Cap at 5 for plotting, flag those > 5
    omega_cap = pmin(wmis_cv, 3),
    capped    = wmis_cv > 3
  )



# Helper to make one panel; show_y controls gene labels
make_panel <- function(v, show_y = FALSE, show_x = FALSE) {
  ggplot(df %>% filter(var == v),
         aes(x = omega_cap, y = gene_name)) +
    geom_vline(xintercept = 1, linetype = "dashed", linewidth = 0.5) +
    geom_point(aes(shape = capped),
               size = 2.2,
               color = variant_colors[[as.character(v)]]) +
    scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17), guide = "none") +
    coord_cartesian(xlim = c(0, 3)) +
    labs(
      title = as.character(v),
      x = if (show_x) "dN/dS (missense, ω)" else NULL,
      y = if (show_y) "Gene" else NULL
    ) +
    theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(size = 12, hjust = 0),
      axis.text.y = if (show_y) element_text(size = 9) else element_blank(),
      axis.title.y = if (show_y) element_text() else element_blank(),
      panel.grid.minor = element_blank()
    )
}

# ---- 4) Build panels (labels only on the first) ----
p_other <- make_panel("Other",        show_y = TRUE)
p_lambda <- make_panel("Lambda",      show_y = FALSE)
p_omi   <- make_panel("Omicron",      show_y = FALSE, show_x = TRUE)
p_coi   <- make_panel("Co-Infection", show_y = FALSE)

# ---- 5) Combine with patchwork ----
gene_panels <- p_other | p_lambda | p_omi | p_coi
gene_panels

final_dn_ds_fig <- p_global + gene_panels + plot_layout(widths = c(1, 3))

pdf("Fig4_dnds.pdf",width = 10, height = 5)
final_dn_ds_fig
dev.off()

final_dn_ds_fig



```

Stadistical tests at gene-level between variants, 
we aim to anwser whether 
```{r statstest}
kruskal.test(as.numeric(wmis_cv) ~ var, data = df)
#pairwise test
FSA::dunnTest(wmis_cv ~ var, data=df, method="bh")

# accounting for gene differences
library(lme4)
m <- lmer(wmis_cv ~ var + (1 | gene_name), data=df)
anova(m)


```

Report table of DN/DS

```{r tables_dnds}
var=df %>% dplyr::select(gene_name,wmis_cv,var)%>% pivot_wider(names_from=var, values_from = wmis_cv)
global_dnds
var
write_csv(var,"dnds_per_gene_var.txt")

a=dnds_oth$sel_cv %>% dplyr::select(gene_name,n_syn,n_mis,wmis_cv,qmis_cv) %>% mutate(var="Other")
b=dnds_lam$sel_cv %>% dplyr::select(gene_name,n_syn,n_mis,wmis_cv,qmis_cv) %>% mutate(var="Lambda")
c=dnds_omi$sel_cv %>% dplyr::select(gene_name,n_syn,n_mis,wmis_cv,qmis_cv) %>% mutate(var="Omicron")
d=dnds_coi$sel_cv %>% dplyr::select(gene_name,n_syn,n_mis,wmis_cv,qmis_cv) %>% mutate(var="Co-Infection")

df_g <- dplyr::bind_rows(a, b, c, d)
df_g=df_g %>% arrange(var,gene_name) 
write_csv(df_g,"dnds_per_gene_var_details.txt")

```


