---
title: "iSNV_COI_V2"
author: "Alex Di Genova"
date: "2026-05-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## iSNV 

iSNV filters from PASS variants called with bcftools

```{r isnv}
library(tidyverse)
library(broom)
library(patchwork)
library(ggbeeswarm)
library(ggpmisc)
library(ggpubr)
library(glue)

# -----------------------------
# Parameters
# -----------------------------
pars <- list(
  af_min_primary = 0.05,
  af_min_sens    = 0.03,
  af_max         = 0.95,
  alt_qual_min   = 30,
  ct_max         = 25,

  # strand support
  alt_rv_min     = 5,   # at least 5 reverse ALT reads
  alt_fw_min     = 5,   # at least 5 forward ALT reads

  # coarse genome-end mask if no primer BED is available
  end_mask_bp    = 100,

  # recurrent hotspot definition
  recurrent_prop_min = 0.10,   # site seen in >=10% of samples
  recurrent_af_max   = 0.10    # but median AF stays low -> suspicious
)

# -----------------------------
# Helper functions
# -----------------------------
calc_strand_bias_p <- function(ref_dp, ref_rv, alt_dp, alt_rv) {
  ref_fw <- ref_dp - ref_rv
  alt_fw <- alt_dp - alt_rv

  # avoid negative weird cases
  if (any(c(ref_fw, ref_rv, alt_fw, alt_rv) < 0)) return(NA_real_)

  m <- matrix(c(ref_fw, ref_rv, alt_fw, alt_rv), nrow = 2, byrow = TRUE)
  fisher.test(m, alternative = "greater")$p.value
}

read_isnv_file <- function(f) {
  read.table(f, header = TRUE, sep = "\t", stringsAsFactors = FALSE) %>%
    mutate(
      Filename = basename(f),
      barcode  = as.integer(stringr::str_extract(basename(f), "(?<=-)\\d+"))
    )
}

# -----------------------------
# Load raw iSNV candidates
# -----------------------------
files_isnv <- list.files("iSNVs", pattern = "_variants.tsv$", full.names = TRUE)

raw_isnv_cache_file <- "analysis/raw_isnv_cache.rds"
raw_isnv_cache_meta <- tibble(
  file = normalizePath(files_isnv),
  size = file.info(files_isnv)$size,
  mtime = file.info(files_isnv)$mtime
)

if (file.exists(raw_isnv_cache_file)) {
  raw_isnv_cache <- readRDS(raw_isnv_cache_file)
} else {
  raw_isnv_cache <- NULL
}

if (
  !is.null(raw_isnv_cache) &&
  identical(raw_isnv_cache$meta, raw_isnv_cache_meta)
) {
  raw_isnv <- raw_isnv_cache$data
} else {
  raw_isnv <- purrr::map_dfr(files_isnv, read_isnv_file) %>%
    filter(PASS=TRUE) %>%
    mutate(
      ALT_FW = ALT_DP - ALT_RV,
      REF_FW = REF_DP - REF_RV,
      strand_p = purrr::pmap_dbl(
        list(REF_DP, REF_RV, ALT_DP, ALT_RV),
        calc_strand_bias_p
      ),
      both_strands_ok = ALT_RV >= pars$alt_rv_min & ALT_FW >= pars$alt_fw_min,
      basic_ok =
        PASS == TRUE &
        ALT_QUAL >= pars$alt_qual_min &
        ALT_FREQ < pars$af_max,
      genome_end = POS <= pars$end_mask_bp | POS >= (29903 - pars$end_mask_bp)
    )

  saveRDS(
    list(meta = raw_isnv_cache_meta, data = raw_isnv),
    raw_isnv_cache_file
  )
}

# -----------------------------
# Load consensus SNVs
# -----------------------------
files_snv <- list.files("SNVs", pattern = "\\.variants.tsv$", full.names = TRUE)

cons_snv <- purrr::map_dfr(files_snv, \(f) {
  read.table(f, header = TRUE, sep = "\t", stringsAsFactors = FALSE) %>%
    mutate(
      Filename = basename(f),
      barcode  = as.integer(stringr::str_extract(basename(f), "(?<=-)\\d+"))
    )
}) %>%
  transmute(barcode, POS, ALT, cons_key = paste(barcode, POS, ALT, sep = "_"))

# -----------------------------
# Sample-level metadata
# df should contain barcode, Ct, etc.
# ff is your Freyja table
# -----------------------------
#coinf_barcodes <- ff %>%
#  filter(percentage > 0.03) %>%
#  group_by(file) %>%
#  summarise(n_variants = n_distinct(variant), .groups = "drop") %>%
#  filter(n_variants > 1) %>%
#  mutate(barcode = as.integer(stringr::str_extract(file, "(?<=-)\\d+"))) %>%
#  pull(barcode)

df=read.table("demographics/all_samples.txt", h=T)
dates_formatted <- as.Date(df$Date, format="%m/%d/%Y")
years <- format(dates_formatted, "%Y")
df$year=years
df=df %>% mutate(Ct=as.integer(Ct))
df$all="all"

# -----------------------------
# loadcoi: Pangolin selected variant plus COI marker
# -----------------------------
all_results <- read.table("analysis/all_results.csv", h = TRUE, sep = ",") %>%
  mutate(
    barcode = as.integer(stringr::str_extract(sampleId, "(?<=_)\\d+")),
    pangolin_variant = case_when(
      stringr::str_detect(variant, "Omicron") ~ "Omicron",
      TRUE ~ variant
    )
  ) %>%
  transmute(
    barcode,
    pangolin_variant,
    mut_ivar,
    pangolin_location = location,
    pangolin_cluster = cluster
  )

coi_summary_file <- "linages_mutations/coinfection_calls_lineage_v3/cohort_coinfection_summary.tsv"

sample_coi <- read.table(coi_summary_file, h = TRUE, sep = "\t") %>%
  mutate(
    barcode = as.integer(stringr::str_extract(sample, "(?<=-)\\d+")),
    co_infection = co_infection_flag %in% c(TRUE, "TRUE", "True", "true", 1, "1"),
    major_label = na_if(stringr::str_trim(major_label), ""),
    minor_label = na_if(stringr::str_trim(minor_label), ""),
    coi_major_lineage = major_label,
    coi_minor_lineage = if_else(co_infection, minor_label, NA_character_),
    coi_lineage_call = if_else(
      co_infection,
      paste(major_label, minor_label, sep = " + "),
      major_label
    )
  ) %>%
  transmute(
    barcode,
    coi_major_lineage,
    coi_major_frac_supported_sites = major_frac_supported_sites,
    coi_major_status = major_status,
    coi_major_n_supported_sites = major_n_supported_sites,
    coi_minor_lineage,
    coi_minor_n_supported_sites = minor_n_supported_sites,
    coi_minor_median_af = minor_median_af,
    coi_confidence_tier = confidence_tier,
    co_infection,
    coi_lineage_call
  )

df <- df %>%
  left_join(all_results, by = "barcode") %>%
  left_join(sample_coi, by = "barcode") %>%
  mutate(
    pangolin_variant = replace_na(pangolin_variant, "Unknown"),
    coi_lineage_call = replace_na(coi_lineage_call, "Unknown"),
    coi_confidence_tier = replace_na(coi_confidence_tier, "none"),
    co_infection = replace_na(co_infection, FALSE),
    coi_major_status = replace_na(coi_major_status, "unassigned"),
    lineage_group = if_else(co_infection, "Co-Infection", pangolin_variant)
  )



sample_qc <- df %>%
  select(barcode, Ct, pangolin_variant, mut_ivar, pangolin_location, pangolin_cluster,
         coi_major_lineage, coi_major_frac_supported_sites, coi_major_status,
         coi_major_n_supported_sites, coi_minor_lineage, coi_minor_n_supported_sites,
         coi_minor_median_af, coi_confidence_tier, co_infection, coi_lineage_call,
         lineage_group) %>%
  mutate(
    sample_ok = Ct <= pars$ct_max
 #   co_infection = barcode %in% coinf_barcodes
  )

# -----------------------------
# Annotate candidate calls
# -----------------------------
raw_isnv <- raw_isnv %>%
  mutate(cons_key = paste(barcode, POS, ALT, sep = "_")) %>%
  left_join(sample_qc, by = "barcode") %>%
  mutate(
    is_consensus_snv = cons_key %in% cons_snv$cons_key,
    site_key = paste(REGION, POS, REF, ALT, sep = "_")
  )

# -----------------------------
# Recurrent suspicious hotspots
# Seen in many samples but with low median AF
# -----------------------------
n_samples <- raw_isnv %>% distinct(barcode) %>% nrow()

hotspots <- raw_isnv %>%
  filter(
    basic_ok,
    ALT_FREQ >= pars$af_min_sens,
    !is.na(sample_ok), sample_ok,
    #!co_infection,
    !is_consensus_snv
  ) %>%
  group_by(site_key, REGION, POS, REF, ALT) %>%
  summarise(
    n_samples_site = n_distinct(barcode),
    prop_samples   = n_samples_site / n_samples,
    median_af      = median(ALT_FREQ, na.rm = TRUE),
    mean_alt_rv    = mean(ALT_RV, na.rm = TRUE),
    mean_alt_fw    = mean(ALT_FW, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    suspicious_hotspot =
      prop_samples >= pars$recurrent_prop_min &
      median_af <= pars$recurrent_af_max
  )

raw_isnv <- raw_isnv %>%
  left_join(
    hotspots %>% select(site_key, suspicious_hotspot),
    by = "site_key"
  ) %>%
  mutate(suspicious_hotspot = replace_na(suspicious_hotspot, FALSE))

# -----------------------------
# Final primary and sensitivity filters
# -----------------------------
isnv_primary <- raw_isnv %>%
  filter(
    basic_ok,
    ALT_FREQ >= pars$af_min_primary,
    !is.na(sample_ok), sample_ok,
   # !co_infection,
    !is_consensus_snv,
    both_strands_ok,
    !genome_end,
    !suspicious_hotspot
  )

isnv_sensitivity <- raw_isnv %>%
  filter(
    basic_ok,
    ALT_FREQ >= pars$af_min_sens,
    !is.na(sample_ok), sample_ok,
   # !co_infection,
    !is_consensus_snv,
    both_strands_ok,
    !genome_end,
    !suspicious_hotspot
  )

```

## iSNV plots

summary of passing variants per filter

```{r pressure, echo=FALSE}
qc_counts <- tibble(
  step = c(
    "Raw",
    "PASS + QUAL",
    "AF >= 3%",
    "Ct <= 28",
    "Not consensus SNV",
    "Both strands",
    "Not genome end",
    "Not recurrent hotspot",
    "Primary AF >= 5%"
  ),
  n = c(
    nrow(raw_isnv),
    sum(raw_isnv$basic_ok, na.rm = TRUE),
    sum(raw_isnv$basic_ok & raw_isnv$ALT_FREQ >= pars$af_min_sens, na.rm = TRUE),
    sum(raw_isnv$basic_ok & raw_isnv$ALT_FREQ >= pars$af_min_sens & raw_isnv$sample_ok, na.rm = TRUE),
    sum(raw_isnv$basic_ok & raw_isnv$ALT_FREQ >= pars$af_min_sens & raw_isnv$sample_ok  & !raw_isnv$is_consensus_snv, na.rm = TRUE),
    sum(raw_isnv$basic_ok & raw_isnv$ALT_FREQ >= pars$af_min_sens & raw_isnv$sample_ok  & !raw_isnv$is_consensus_snv & raw_isnv$both_strands_ok, na.rm = TRUE),
    sum(raw_isnv$basic_ok & raw_isnv$ALT_FREQ >= pars$af_min_sens & raw_isnv$sample_ok  & !raw_isnv$is_consensus_snv & raw_isnv$both_strands_ok & !raw_isnv$genome_end, na.rm = TRUE),
    sum(raw_isnv$basic_ok & raw_isnv$ALT_FREQ >= pars$af_min_sens & raw_isnv$sample_ok  & !raw_isnv$is_consensus_snv & raw_isnv$both_strands_ok & !raw_isnv$genome_end & !raw_isnv$suspicious_hotspot, na.rm = TRUE),
    nrow(isnv_primary)
  )
)

ggplot(qc_counts, aes(x = step, y = log(n))) +
  geom_col() +
  coord_flip() +
  theme_minimal() +
  labs(x = NULL, y = "Number of candidate log(iSNVs)")
```

```{r qc_isnv_sensiviti}
qc_counts <- tibble(
  step = c(
    "Raw",
    "PASS + QUAL",
    "AF >= 3%",
    "Ct <= 28",
    "Not consensus SNV",
    "Both strands",
    "Not genome end",
    "Not recurrent hotspot",
    "Primary AF >= 3%"
  ),
  n = c(
    nrow(raw_isnv),
    sum(raw_isnv$basic_ok, na.rm = TRUE),
    sum(raw_isnv$basic_ok & raw_isnv$ALT_FREQ >= pars$af_min_sens, na.rm = TRUE),
    sum(raw_isnv$basic_ok & raw_isnv$ALT_FREQ >= pars$af_min_sens & raw_isnv$sample_ok, na.rm = TRUE),
    sum(raw_isnv$basic_ok & raw_isnv$ALT_FREQ >= pars$af_min_sens & raw_isnv$sample_ok  & !raw_isnv$is_consensus_snv, na.rm = TRUE),
    sum(raw_isnv$basic_ok & raw_isnv$ALT_FREQ >= pars$af_min_sens & raw_isnv$sample_ok  & !raw_isnv$is_consensus_snv & raw_isnv$both_strands_ok, na.rm = TRUE),
    sum(raw_isnv$basic_ok & raw_isnv$ALT_FREQ >= pars$af_min_sens & raw_isnv$sample_ok  & !raw_isnv$is_consensus_snv & raw_isnv$both_strands_ok & !raw_isnv$genome_end, na.rm = TRUE),
    sum(raw_isnv$basic_ok & raw_isnv$ALT_FREQ >= pars$af_min_sens & raw_isnv$sample_ok  & !raw_isnv$is_consensus_snv & raw_isnv$both_strands_ok & !raw_isnv$genome_end & !raw_isnv$suspicious_hotspot, na.rm = TRUE),
    nrow(isnv_sensitivity)
  )
)

ggplot(qc_counts, aes(x = step, y = log(n))) +
  geom_col() +
  coord_flip() +
  theme_minimal() +
  labs(x = NULL, y = "Number of candidate log(iSNVs)")

```



```{r}

# hotspots %>%
#   ggplot(aes(x = prop_samples, y = median_af)) +
#   geom_point(alpha = 0.6) +
#   theme_minimal() +
#   labs(x = "Fraction of samples with site", y = "Median ALT frequency")

```

## Primary iSNV variable diagnostics

These plots inspect the retained `isnv_primary` calls directly. They are useful for checking
whether the final calls have plausible allele frequencies, read depths, strand support,
quality values, genomic distribution, and mutation spectra.

```{r isnv-primary-basic-plots, fig.width=12, fig.height=10}
isnv_primary_plot <- isnv_primary %>%
  mutate(
    mutation = paste0(REF, ">", ALT),
    strand_balance = ALT_FW / ALT_DP,
    log_total_dp = log10(TOTAL_DP),
    log_alt_dp = log10(ALT_DP)
  )

p_af <- ggplot(isnv_primary_plot, aes(x = ALT_FREQ)) +
  geom_histogram(binwidth = 0.025, fill = "grey35", color = "white") +
  theme_minimal() +
  labs(x = "ALT frequency", y = "Primary iSNV count")

p_depth <- ggplot(isnv_primary_plot, aes(x = TOTAL_DP, y = ALT_DP)) +
  geom_point(aes(color = ALT_FREQ), alpha = 0.65, size = 1.7) +
  scale_x_log10() +
  scale_y_log10() +
  theme_minimal() +
  labs(x = "Total depth", y = "ALT depth", color = "ALT frequency")

p_strand <- ggplot(isnv_primary_plot, aes(x = strand_balance)) +
  geom_histogram(binwidth = 0.05, fill = "grey35", color = "white") +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "grey45") +
  theme_minimal() +
  labs(x = "ALT forward-read fraction", y = "Primary iSNV count")

p_qual <- ggplot(isnv_primary_plot, aes(x = ALT_QUAL)) +
  geom_bar(fill = "grey35") +
  theme_minimal() +
  labs(x = "ALT quality", y = "Primary iSNV count")

p_pos <- ggplot(isnv_primary_plot, aes(x = POS)) +
  geom_histogram(binwidth = 500, fill = "grey35", color = "white") +
  theme_minimal() +
  labs(x = "Genome position", y = "Primary iSNV count")

p_mut <- ggplot(isnv_primary_plot, aes(x = mutation, fill = mutation)) +
  geom_bar() +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "Mutation type", y = "Primary iSNV count")

(p_af | p_depth) / (p_strand | p_qual) / (p_pos | p_mut)
```

```{r isnv-sensitivity-sample-covariates, fig.width=11, fig.height=7}
p_af_ct <- ggplot(isnv_primary_plot, aes(x = Ct, y = ALT_FREQ)) +
  geom_point(aes(color = mutation), alpha = 0.55, size = 1.6) +
  geom_smooth(method = "loess", se = TRUE, color = "black") +
  theme_minimal() +
  labs(x = "Ct", y = "ALT frequency", color = "Mutation")

p_dp_ct <- ggplot(isnv_primary_plot, aes(x = Ct, y = TOTAL_DP)) +
  geom_point(aes(color = mutation), alpha = 0.55, size = 1.6) +
  geom_smooth(method = "loess", se = TRUE, color = "black") +
  scale_y_log10() +
  theme_minimal() +
  labs(x = "Ct", y = "Total depth", color = "Mutation")

p_af_barcode <- isnv_primary_plot %>%
  group_by(barcode) %>%
  summarise(
    n_isnv = n(),
    median_af = median(ALT_FREQ, na.rm = TRUE),
    Ct = first(Ct),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = reorder(factor(barcode), n_isnv), y = n_isnv, fill = median_af)) +
  geom_col() +
  coord_flip() +
  theme_minimal() +
  labs(x = "Barcode", y = "Primary iSNVs", fill = "Median AF")

p_af_ct / p_dp_ct / p_af_barcode
```

## Sensivity iSNV variable diagnostics

These plots inspect the retained `isnv_sensivity` calls directly. 

```{r isnv-sensivity-basic-plots, fig.width=12, fig.height=10}
isnv_sensitivity_plot <- isnv_sensitivity %>%
  mutate(
    mutation = paste0(REF, ">", ALT),
    strand_balance = ALT_FW / ALT_DP,
    log_total_dp = log10(TOTAL_DP),
    log_alt_dp = log10(ALT_DP)
  )

p_af <- ggplot(isnv_sensitivity_plot, aes(x = ALT_FREQ)) +
  geom_histogram(binwidth = 0.025, fill = "grey35", color = "white") +
  theme_minimal() +
  labs(x = "ALT frequency", y = "Sensivity iSNV count")

p_depth <- ggplot(isnv_sensitivity_plot, aes(x = TOTAL_DP, y = ALT_DP)) +
  geom_point(aes(color = ALT_FREQ), alpha = 0.65, size = 1.7) +
  scale_x_log10() +
  scale_y_log10() +
  theme_minimal() +
  labs(x = "Total depth", y = "ALT depth", color = "ALT frequency")

p_strand <- ggplot(isnv_sensitivity_plot, aes(x = strand_balance)) +
  geom_histogram(binwidth = 0.05, fill = "grey35", color = "white") +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "grey45") +
  theme_minimal() +
  labs(x = "ALT forward-read fraction", y = "Primary iSNV count")

p_qual <- ggplot(isnv_sensitivity_plot, aes(x = ALT_QUAL)) +
  geom_bar(fill = "grey35") +
  theme_minimal() +
  labs(x = "ALT quality", y = "Sensivity iSNV count")

p_pos <- ggplot(isnv_sensitivity_plot, aes(x = POS)) +
  geom_histogram(binwidth = 500, fill = "grey35", color = "white") +
  theme_minimal() +
  labs(x = "Genome position", y = "Primary iSNV count")

p_mut <- ggplot(isnv_sensitivity_plot, aes(x = mutation, fill = mutation)) +
  geom_bar() +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "Mutation type", y = "Sensivity iSNV count")

(p_af | p_depth) / (p_strand | p_qual) / (p_pos | p_mut)
```

```{r isnv-primary-sample-covariates, fig.width=11, fig.height=7}
p_af_ct <- ggplot(isnv_sensitivity_plot, aes(x = Ct, y = ALT_FREQ)) +
  geom_point(aes(color = mutation), alpha = 0.55, size = 1.6) +
  geom_smooth(method = "loess", se = TRUE, color = "black") +
  theme_minimal() +
  labs(x = "Ct", y = "ALT frequency", color = "Mutation")

p_dp_ct <- ggplot(isnv_sensitivity_plot, aes(x = Ct, y = TOTAL_DP)) +
  geom_point(aes(color = mutation), alpha = 0.55, size = 1.6) +
  geom_smooth(method = "loess", se = TRUE, color = "black") +
  scale_y_log10() +
  theme_minimal() +
  labs(x = "Ct", y = "Total depth", color = "Mutation")

p_af_barcode <- isnv_sensitivity_plot %>%
  group_by(barcode) %>%
  summarise(
    n_isnv = n(),
    median_af = median(ALT_FREQ, na.rm = TRUE),
    Ct = first(Ct),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = reorder(factor(barcode), n_isnv), y = n_isnv, fill = median_af)) +
  geom_col() +
  coord_flip() +
  theme_minimal() +
  labs(x = "Barcode", y = "Primary iSNVs", fill = "Median AF")

p_af_ct / p_dp_ct / p_af_barcode
```


## Per-sample iSNV burden

These diagnostics use the number of primary iSNVs per sample as the response. Samples with
no retained primary iSNVs are kept as zero counts, which is required for an unbiased
count-regression analysis.

```{r isnv-sample-burden}
isnv_counts <- isnv_primary %>%
  group_by(barcode) %>%
  summarise(
    n_isnv = n(),
    mean_alt_freq = mean(ALT_FREQ, na.rm = TRUE),
    median_alt_freq = median(ALT_FREQ, na.rm = TRUE),
    mean_total_dp = mean(TOTAL_DP, na.rm = TRUE),
    median_total_dp = median(TOTAL_DP, na.rm = TRUE),
    .groups = "drop"
  )

isnv_sample <- df %>%
  mutate(
    barcode = as.integer(barcode),
    Date = as.Date(Date, format = "%m/%d/%Y"),
    collection_day = as.numeric(Date - min(Date, na.rm = TRUE)),
    Sex = factor(Sex),
    Region = factor(Region),
    pangolin_variant = factor(pangolin_variant),
    co_infection = factor(co_infection, levels = c(FALSE, TRUE), labels = c("No", "Yes")),
    lineage_group = factor(lineage_group),
    sample_ok = Ct <= pars$ct_max
  ) %>%
  left_join(isnv_counts, by = "barcode") %>%
  mutate(
    n_isnv = replace_na(n_isnv, 0L),
    any_isnv = n_isnv > 0,
    log_ct = log(Ct),
    mean_alt_freq = if_else(n_isnv == 0, NA_real_, mean_alt_freq),
    median_alt_freq = if_else(n_isnv == 0, NA_real_, median_alt_freq)
  )

isnv_sample_qc <- isnv_sample %>%
  filter(sample_ok)

add_coi_medium_plot_group <- function(dat) {
  base_dat <- dat %>%
    mutate(lineage_plot_group = as.character(lineage_group))

  medium_coi_dat <- base_dat %>%
    filter(lineage_group == "Co-Infection", coi_confidence_tier == "medium") %>%
    mutate(lineage_plot_group = "Co-Infection medium")

  plot_dat <- bind_rows(base_dat, medium_coi_dat)
  preferred_levels <- c("Other", "Lambda", "Omicron", "Co-Infection", "Co-Infection medium")
  extra_levels <- setdiff(sort(unique(plot_dat$lineage_plot_group)), preferred_levels)

  plot_dat %>%
    mutate(
      lineage_plot_group = factor(
        lineage_plot_group,
        levels = c(preferred_levels, extra_levels)
      )
    )
}

isnv_sample_qc %>%
  summarise(
    n_samples = n(),
    samples_with_isnv = sum(any_isnv),
    median_isnv = median(n_isnv),
    mean_isnv = mean(n_isnv),
    variance_isnv = var(n_isnv),
    max_isnv = max(n_isnv)
  )
```

```{r isnv-count-diagnostics, fig.width=11, fig.height=8}
p_count_hist <- ggplot(isnv_sample_qc, aes(x = n_isnv)) +
  geom_histogram(binwidth = 1, boundary = -0.5, fill = "grey35", color = "white") +
  theme_minimal() +
  labs(x = "Primary iSNVs per sample", y = "Samples")

p_ct <- add_coi_medium_plot_group(isnv_sample_qc) %>%
  ggplot(aes(x = Ct, y = n_isnv)) +
  geom_point(aes(color = lineage_plot_group), size = 2, alpha = 0.8) +
  geom_smooth(data = isnv_sample_qc, method = "glm", method.args = list(family = "poisson"), se = TRUE, color = "black") +
  theme_minimal() +
  labs(x = "Ct", y = "Primary iSNVs per sample", color = "Variant group")

p_age <- ggplot(isnv_sample_qc, aes(x = Age, y = n_isnv)) +
  geom_point(aes(color = Sex), size = 2, alpha = 0.8) +
  geom_smooth(method = "glm", method.args = list(family = "poisson"), se = TRUE, color = "black") +
  theme_minimal() +
  labs(x = "Age", y = "Primary iSNVs per sample", color = "Sex")

p_date <- add_coi_medium_plot_group(isnv_sample_qc) %>%
  ggplot(aes(x = Date, y = n_isnv)) +
  geom_point(aes(color = lineage_plot_group), size = 2, alpha = 0.8) +
  geom_smooth(data = isnv_sample_qc, method = "glm", method.args = list(family = "poisson"), se = TRUE, color = "black") +
  theme_minimal() +
  labs(x = "Collection date", y = "Primary iSNVs per sample", color = "Variant group")

p_group <- add_coi_medium_plot_group(isnv_sample_qc) %>%
  pivot_longer(c(Sex, lineage_plot_group), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value, y = n_isnv, fill = value)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.45) +
  geom_jitter(width = 0.12, height = 0.05, alpha = 0.65, size = 1.8) +
  facet_wrap(~ variable, scales = "free_x") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = NULL, y = "Primary iSNVs per sample")

(p_count_hist | p_ct) / (p_age | p_date) / p_group
```

## Comparison iSNV vs SNV

This reproduces the SNV/iSNV comparison from `MFig_Covid.Rmd` using the cleaned
`isnv_primary` calls and keeping zero-count samples in both burdens.

```{r cmpsnvs_clean, fig.width=12, fig.height=7}
MutSNV <- cons_snv %>%
  count(barcode, name = "n_snv") %>%
  right_join(df, by = "barcode") %>%
  mutate(
    n_snv = replace_na(n_snv, 0L),
    Date = as.Date(Date, format = "%m/%d/%Y")
  ) %>%
  filter(Ct <= pars$ct_max)

iSNV_date_data <- isnv_sample_qc %>%
  mutate(Date = as.Date(Date, format = "%m/%d/%Y"))

MutSNV_plot <- add_coi_medium_plot_group(MutSNV)
iSNV_date_plot <- add_coi_medium_plot_group(iSNV_date_data)

p_snv_sex <- MutSNV_plot %>%
  ggplot(aes(x = Sex, y = n_snv, color = lineage_plot_group)) +
  geom_quasirandom(width = 0.18, alpha = 0.85) +
  facet_wrap(~ year) +
  labs(x = "Sexo", y = "SNV", color = "Variant group", title = "SNVs per year") +
  theme_light() +
  theme(legend.position = "none")

p_isnv_sex <- iSNV_date_plot %>%
  ggplot(aes(x = Sex, y = n_isnv, color = lineage_plot_group)) +
  geom_quasirandom(width = 0.18, alpha = 0.85) +
  facet_wrap(~ year) +
  labs(x = "Sexo", y = "iSNV", color = "Variant group", title = "iSNVs per year") +
  theme_light() +
  theme(legend.position = "none")

p_snv_date <- MutSNV_plot %>%
  ggplot(aes(x = Date, y = n_snv)) +
  geom_point(aes(color = lineage_plot_group), size = 2, alpha = 0.9) +
  geom_smooth(data = MutSNV, method = "lm", se = TRUE) +
  stat_poly_eq(
    data = MutSNV,
    aes(label = paste(after_stat(eq.label), after_stat(rr.label), sep = "~~~")),
    formula = y ~ x,
    parse = TRUE
  ) +
  scale_x_date(date_breaks = "4 months", date_labels = "%b/%Y") +
  labs(title = "SNVs vs Date", x = "Date", y = "SNV") +
  theme_light() +
  theme(legend.position = "none")

p_isnv_date <- iSNV_date_plot %>%
  ggplot(aes(x = Date, y = n_isnv)) +
  geom_point(aes(color = lineage_plot_group), size = 2, alpha = 0.9) +
  geom_smooth(data = iSNV_date_data, method = "lm", se = TRUE) +
  stat_poly_eq(
    data = iSNV_date_data,
    aes(label = paste(after_stat(eq.label), after_stat(rr.label), sep = "~~~")),
    formula = y ~ x,
    parse = TRUE
  ) +
  scale_x_date(date_breaks = "4 months", date_labels = "%b/%Y") +
  labs(title = "iSNVs vs Date", x = "Date", y = "iSNV") +
  theme_light() +
  theme(legend.position = "bottom")

(p_snv_sex / p_isnv_sex) | (p_snv_date / p_isnv_date)
```

```{r isnv_snv_date_lm}
snv_date_lm <- lm(n_snv ~ Date, data = MutSNV)
isnv_date_lm <- lm(n_isnv ~ Date, data = iSNV_date_data)

summary(snv_date_lm)
summary(isnv_date_lm)
```

## Linear Regression Model

This linear model using iSNV burden and Pangolin/COI annotations from this analysis.

```{r isnv_linear_model}
coverage_file <- if (file.exists("resume_samtools_coverage.txt")) {
  "resume_samtools_coverage.txt"
} else {
  "analysis/resume_samtools_coverage.txt"
}

coverage_qc <- read.table(coverage_file, h = TRUE) %>%
  mutate(
    barcode = as.integer(stringr::str_extract(sname, "(?<=-)\\d+")),
    logc = log10(meandepth)
  ) %>%
  select(barcode, meandepth, coverage, logc)

variant_min_n <- 5

valid_lineage_groups <- isnv_sample_qc %>%
  count(lineage_group, name = "n_samples_group") %>%
  filter(n_samples_group >= variant_min_n) %>%
  pull(lineage_group)

isnv_lm_data <- isnv_sample_qc %>%
  filter(lineage_group %in% valid_lineage_groups) %>%
  left_join(coverage_qc, by = "barcode") %>%
  mutate(
    age2 = case_when(
      Age < 30 ~ "young",
      Age >= 30 & Age < 50 ~ "adult",
      TRUE ~ "older"
    ),
    age2 = relevel(factor(age2), ref = "adult"),
    Sex = factor(Sex),
    lineage_group = relevel(droplevels(factor(lineage_group)), ref = "Other"),
    n = n_isnv
  ) %>%
  drop_na(n, age2, Sex, mut_ivar, lineage_group, Ct, logc)

isnv_lm_data %>%
  count(lineage_group, name = "n_samples_group") %>%
  arrange(desc(n_samples_group))

iSNVlm <- lm(n ~ age2 + Sex + mut_ivar + lineage_group + Ct + logc, data = isnv_lm_data)
summary(iSNVlm)
```

```{r lmmedian_test, fig.width=10, fig.height=7}
fit <- iSNVlm

coef_df <- tidy(fit, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term_label = case_when(
      term == "age2older" ~ "Age: older (vs adult)",
      term == "age2young" ~ "Age: young (vs adult)",
      term == "SexM" ~ "Sex: male (vs female)",
      term == "Ct" ~ "PCR Ct value",
      term == "logc" ~ "Average genome coverage (log10)",
      term == "mut_ivar" ~ "iVar mutation load",
      str_starts(term, "lineage_group") ~ str_replace(term, "lineage_group", "Variant: "),
      TRUE ~ term
    ),
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      p.value < 0.1 ~ ".",
      TRUE ~ ""
    ),
    term_label = reorder(term_label, estimate)
  )

gl <- glance(fit)
subtitle_txt <- glue(
  "Adj. R2 = {round(gl$adj.r.squared, 3)}; F({gl$df}, {gl$df.residual}) = ",
  "{round(gl$statistic, 2)}, p = {signif(gl$p.value, 3)}"
)

p_model <- ggplot(coef_df, aes(x = term_label, y = estimate)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2, linewidth = 0.6) +
  geom_point(size = 2.3) +
  geom_text(aes(label = sig, y = conf.high), vjust = -0.6, size = 4) +
  coord_flip() +
  labs(
    title = "A",
    subtitle = subtitle_txt,
    x = NULL,
    y = "Effect on iSNV count"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(size = 16, hjust = 0, face = "plain"),
    plot.subtitle = element_text(color = "grey20", size = 11),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 10),
    axis.title.y = element_text(size = 11)
  )

sex_comparisons <- list(c("F", "M"))
y_upper <- max(isnv_lm_data$n, na.rm = TRUE) * 1.12

p_sex <- ggboxplot(
  isnv_lm_data, x = "Sex", y = "n",
  color = "Sex", palette = "jco",
  add = "jitter"
) +
  stat_compare_means(
    comparisons = sex_comparisons,
    method = "wilcox.test",
    exact = FALSE,
    label = "p.format"
  ) +
  coord_cartesian(ylim = c(0, y_upper)) +
  labs(title = "B", x = "Sex", y = NULL) +
  theme_pubr(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 16, hjust = -0.1, face = "plain")
  )

mvars <- c("Lambda", "Omicron", "Other", "Co-Infection", "Co-Infection medium")
isnv_lm_variant_data <- isnv_lm_data %>%
  add_coi_medium_plot_group() %>%
  filter(lineage_plot_group %in% mvars) %>%
  mutate(lineage_plot_group = factor(lineage_plot_group, levels = mvars))

pairwise_comparisons <- list(
  c("Other", "Lambda"),
  c("Omicron", "Other"),
  c("Co-Infection", "Omicron"),
  c("Co-Infection", "Other"),
  c("Co-Infection", "Lambda"),
  c("Co-Infection medium", "Omicron"),
  c("Co-Infection medium", "Other"),
  c("Co-Infection medium", "Lambda")
)

pairwise_comparisons <- pairwise_comparisons[
  purrr::map_lgl(pairwise_comparisons, ~ all(.x %in% unique(isnv_lm_variant_data$lineage_plot_group)))
]

variant_colors <- c(
  "Lambda" = "#8E7CC3",
  "Omicron" = "#B45F06",
  "Other" = "#38761D",
  "Co-Infection" = "#333333",
  "Co-Infection medium" = "#CC0000"
)

p_variant <- ggboxplot(
  isnv_lm_variant_data, x = "lineage_plot_group", y = "n",
  color = "lineage_plot_group", palette = variant_colors,
  add = "jitter"
) +
  stat_compare_means(comparisons = pairwise_comparisons, exact = FALSE, label = "p.format") +
  stat_compare_means(label.y = y_upper * 0.98, method = "kruskal.test", label = "p.format") +
  coord_cartesian(ylim = c(0, y_upper)) +
  labs(title = "C", x = "SARS-CoV-2 variant / COI group", y = NULL) +
  theme_pubr(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 10, angle = 20, hjust = 1),
    axis.title.x = element_text(size = 11, margin = margin(t = 10)),
    plot.title = element_text(size = 16, hjust = -0.1, face = "plain")
  )

y_shared <- ggplot() +
  annotate("text", x = 0, y = 0, label = "Number of iSNVs", angle = 90, size = 4) +
  theme_void() +
  theme(plot.margin = margin(r = 2))

right_block <- (p_sex / p_variant) + plot_layout(heights = c(1, 1))
right_block_with_ylabel <- (y_shared | right_block) + plot_layout(widths = c(0.06, 1))

final_fig <- (p_model | right_block_with_ylabel) + plot_layout(widths = c(1.05, 1.35))
final_fig
```

```{r isnv-count-models}
count_model_data <- isnv_sample_qc %>%
  filter(lineage_group %in% valid_lineage_groups) %>%
  select(n_isnv, Ct, Age, Sex, lineage_group, collection_day) %>%
  drop_na() %>%
  mutate(lineage_group = relevel(droplevels(factor(lineage_group)), ref = "Other"))

poisson_fit <- glm(
  n_isnv ~ Ct + Age + Sex + lineage_group + collection_day,
  family = poisson(link = "log"),
  data = count_model_data
)

dispersion_ratio <- sum(residuals(poisson_fit, type = "pearson")^2) /
  poisson_fit$df.residual

nb_fit <- MASS::glm.nb(
  n_isnv ~ Ct + Age + Sex + lineage_group + collection_day,
  data = count_model_data
)

model_comparison <- tibble(
  model = c("Poisson", "Negative binomial"),
  AIC = c(AIC(poisson_fit), AIC(nb_fit)),
  dispersion_ratio = c(dispersion_ratio, NA_real_)
)

model_comparison

selected_fit <- if (dispersion_ratio > 1.2) nb_fit else poisson_fit

selected_model_name <- if (identical(selected_fit, nb_fit)) {
  "Negative binomial"
} else {
  "Poisson"
}

selected_model_name
```

```{r isnv-model-effects, fig.width=8, fig.height=5}
effect_table <- tidy(selected_fit, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term = recode(
      term,
      Ct = "Ct",
      Age = "Age",
      collection_day = "Collection day"
    ),
    direction = case_when(
      conf.low > 1 ~ "Positive",
      conf.high < 1 ~ "Negative",
      TRUE ~ "Uncertain"
    )
  ) %>%
  arrange(estimate)

effect_table

ggplot(effect_table, aes(x = estimate, y = reorder(term, estimate), color = direction)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey45") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.15) +
  geom_point(size = 2.5) +
  scale_x_log10() +
  theme_minimal() +
  labs(
    x = "Rate ratio for primary iSNV count",
    y = NULL,
    color = "Direction",
    title = paste("Predictors of iSNV burden:", selected_model_name, "model")
  )
```

```{r isnv-model-diagnostics, fig.width=10, fig.height=4}
diagnostic_data <- count_model_data %>%
  mutate(
    fitted = fitted(selected_fit),
    pearson_resid = residuals(selected_fit, type = "pearson"),
    response_resid = residuals(selected_fit, type = "response")
  )

p_fit <- ggplot(diagnostic_data, aes(x = fitted, y = n_isnv)) +
  geom_point(alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey45") +
  theme_minimal() +
  labs(x = "Fitted iSNV count", y = "Observed iSNV count")

p_resid <- ggplot(diagnostic_data, aes(x = fitted, y = pearson_resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey45") +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Fitted iSNV count", y = "Pearson residual")

p_fit | p_resid
```

## Primary vs Sensitivity iSNV Set Comparison

This section repeats the same sample-level diagnostics and regressions for the
`isnv_primary` and `isnv_sensitivity` call sets. Variant groups with fewer than
`variant_min_n` samples are removed from model/testing datasets only.

```{r compare-primary-sensitivity-build}
build_isnv_sample_set <- function(isnv_calls, set_name, sample_df = df) {
  set_counts <- isnv_calls %>%
    group_by(barcode) %>%
    summarise(
      n_isnv = n(),
      mean_alt_freq = mean(ALT_FREQ, na.rm = TRUE),
      median_alt_freq = median(ALT_FREQ, na.rm = TRUE),
      mean_total_dp = mean(TOTAL_DP, na.rm = TRUE),
      median_total_dp = median(TOTAL_DP, na.rm = TRUE),
      .groups = "drop"
    )

  sample_df %>%
    mutate(
      barcode = as.integer(barcode),
      Date = as.Date(Date, format = "%m/%d/%Y"),
      collection_day = as.numeric(Date - min(Date, na.rm = TRUE)),
      Sex = factor(Sex),
      Region = factor(Region),
      lineage_group = factor(lineage_group),
      sample_ok = Ct <= pars$ct_max
    ) %>%
    left_join(set_counts, by = "barcode") %>%
    mutate(
      isnv_set = set_name,
      n_isnv = replace_na(n_isnv, 0L),
      any_isnv = n_isnv > 0,
      mean_alt_freq = if_else(n_isnv == 0, NA_real_, mean_alt_freq),
      median_alt_freq = if_else(n_isnv == 0, NA_real_, median_alt_freq)
    )
}

isnv_set_sample_qc <- bind_rows(
  build_isnv_sample_set(isnv_primary, "Primary"),
  build_isnv_sample_set(isnv_sensitivity, "Sensitivity")
) %>%
  filter(sample_ok)

isnv_set_summary <- isnv_set_sample_qc %>%
  group_by(isnv_set) %>%
  summarise(
    n_samples = n(),
    samples_with_isnv = sum(any_isnv),
    median_isnv = median(n_isnv),
    mean_isnv = mean(n_isnv),
    variance_isnv = var(n_isnv),
    max_isnv = max(n_isnv),
    .groups = "drop"
  )

isnv_set_summary
```

```{r compare-primary-sensitivity-diagnostics, fig.width=12, fig.height=9}
p_set_hist <- ggplot(isnv_set_sample_qc, aes(x = n_isnv)) +
  geom_histogram(binwidth = 1, boundary = -0.5, fill = "grey35", color = "white") +
  facet_wrap(~ isnv_set, scales = "free_y") +
  theme_minimal() +
  labs(x = "iSNVs per sample", y = "Samples")

p_set_ct <- add_coi_medium_plot_group(isnv_set_sample_qc) %>%
  ggplot(aes(x = Ct, y = n_isnv)) +
  geom_point(aes(color = lineage_plot_group), alpha = 0.75, size = 1.8) +
  geom_smooth(data = isnv_set_sample_qc, method = "glm", method.args = list(family = "poisson"), se = TRUE, color = "black") +
  facet_wrap(~ isnv_set, scales = "free_y") +
  theme_minimal() +
  labs(x = "Ct", y = "iSNVs per sample", color = "Variant group")

p_set_date <- add_coi_medium_plot_group(isnv_set_sample_qc) %>%
  ggplot(aes(x = Date, y = n_isnv)) +
  geom_point(aes(color = lineage_plot_group), alpha = 0.85, size = 1.8) +
  geom_smooth(data = isnv_set_sample_qc, method = "lm", se = TRUE, color = "black") +
  facet_wrap(~ isnv_set, scales = "free_y") +
  theme_minimal() +
  labs(x = "Collection date", y = "iSNVs per sample", color = "Variant group")

p_set_variant <- add_coi_medium_plot_group(isnv_set_sample_qc) %>%
  group_by(isnv_set, lineage_plot_group) %>%
  filter(n() >= variant_min_n) %>%
  ungroup() %>%
  ggplot(aes(x = lineage_plot_group, y = n_isnv, fill = lineage_plot_group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.45) +
  geom_jitter(width = 0.12, height = 0.05, alpha = 0.65, size = 1.7) +
  facet_wrap(~ isnv_set, scales = "free_y") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1)) +
  labs(x = "Variant group (n >= 5)", y = "iSNVs per sample")

(p_set_hist | p_set_ct) / (p_set_date | p_set_variant)
```

```{r compare-primary-sensitivity-tests}
isnv_set_model_data <- isnv_set_sample_qc %>%
  group_by(isnv_set, lineage_group) %>%
  filter(n() >= variant_min_n) %>%
  ungroup() %>%
  left_join(coverage_qc, by = "barcode") %>%
  mutate(
    age2 = case_when(
      Age < 30 ~ "young",
      Age >= 30 & Age < 50 ~ "adult",
      TRUE ~ "older"
    ),
    age2 = relevel(factor(age2), ref = "adult"),
    lineage_group = factor(lineage_group),
    lineage_group = if ("Other" %in% levels(lineage_group)) {
      relevel(lineage_group, ref = "Other")
    } else {
      lineage_group
    },
    n = n_isnv
  ) %>%
  drop_na(n, age2, Sex, mut_ivar, lineage_group, Ct, logc, collection_day)

variant_group_counts_by_set <- isnv_set_model_data %>%
  count(isnv_set, lineage_group, name = "n_samples_group") %>%
  arrange(isnv_set, desc(n_samples_group))

variant_group_counts_by_set

sex_wilcox_by_set <- isnv_set_model_data %>%
  group_by(isnv_set) %>%
  summarise(
    p_value = if (n_distinct(Sex) == 2) {
      wilcox.test(n ~ Sex, exact = FALSE)$p.value
    } else {
      NA_real_
    },
    .groups = "drop"
  )

variant_kruskal_by_set <- isnv_set_model_data %>%
  group_by(isnv_set) %>%
  summarise(
    p_value = if (n_distinct(lineage_group) > 1) {
      kruskal.test(n ~ lineage_group)$p.value
    } else {
      NA_real_
    },
    .groups = "drop"
  )

sex_wilcox_by_set
variant_kruskal_by_set
```

```{r compare-primary-sensitivity-regressions}
fit_set_models <- function(dat) {
  dat <- dat %>%
    mutate(
      lineage_group = droplevels(lineage_group),
      lineage_group = if ("Other" %in% levels(lineage_group)) {
        relevel(lineage_group, ref = "Other")
      } else {
        lineage_group
      }
    )

  poisson_fit <- glm(
    n ~ Ct + Age + Sex + lineage_group + collection_day,
    family = poisson(link = "log"),
    data = dat
  )

  dispersion_ratio <- sum(residuals(poisson_fit, type = "pearson")^2) /
    poisson_fit$df.residual

  nb_fit <- tryCatch(
    MASS::glm.nb(
      n ~ Ct + Age + Sex + lineage_group + collection_day,
      data = dat
    ),
    error = function(e) NULL
  )

  selected_count_fit <- if (!is.null(nb_fit) && dispersion_ratio > 1.2) nb_fit else poisson_fit
  selected_count_model <- if (!is.null(nb_fit) && dispersion_ratio > 1.2) "Negative binomial" else "Poisson"

  lm_fit <- lm(n ~ age2 + Sex + mut_ivar + lineage_group + Ct + logc, data = dat)

  list(
    poisson_fit = poisson_fit,
    nb_fit = nb_fit,
    selected_count_fit = selected_count_fit,
    selected_count_model = selected_count_model,
    dispersion_ratio = dispersion_ratio,
    lm_fit = lm_fit
  )
}

set_model_fits <- isnv_set_model_data %>%
  group_by(isnv_set) %>%
  group_split() %>%
  set_names(purrr::map_chr(., ~ unique(.x$isnv_set))) %>%
  purrr::map(fit_set_models)

count_model_comparison_by_set <- purrr::imap_dfr(set_model_fits, \(fit, set_name) {
  tibble(
    isnv_set = set_name,
    selected_count_model = fit$selected_count_model,
    poisson_aic = AIC(fit$poisson_fit),
    nb_aic = if (!is.null(fit$nb_fit)) AIC(fit$nb_fit) else NA_real_,
    dispersion_ratio = fit$dispersion_ratio
  )
})

count_effects_by_set <- purrr::imap_dfr(set_model_fits, \(fit, set_name) {
  tidy(fit$selected_count_fit, conf.int = TRUE, exponentiate = TRUE) %>%
    filter(term != "(Intercept)") %>%
    mutate(
      isnv_set = set_name,
      selected_count_model = fit$selected_count_model,
      direction = case_when(
        conf.low > 1 ~ "Positive",
        conf.high < 1 ~ "Negative",
        TRUE ~ "Uncertain"
      )
    )
})

lm_effects_by_set <- purrr::imap_dfr(set_model_fits, \(fit, set_name) {
  tidy(fit$lm_fit, conf.int = TRUE) %>%
    filter(term != "(Intercept)") %>%
    mutate(
      isnv_set = set_name,
      direction = case_when(
        conf.low > 0 ~ "Positive",
        conf.high < 0 ~ "Negative",
        TRUE ~ "Uncertain"
      )
    )
})

count_model_comparison_by_set
count_effects_by_set
lm_effects_by_set
```

```{r compare-primary-sensitivity-regression-plots, fig.width=12, fig.height=9}
p_count_effects_sets <- count_effects_by_set %>%
  mutate(term_label = case_when(
    term == "Ct" ~ "Ct",
    term == "Age" ~ "Age",
    term == "collection_day" ~ "Collection day",
    str_starts(term, "lineage_group") ~ str_replace(term, "lineage_group", "Variant: "),
    TRUE ~ term
  )) %>%
  ggplot(aes(x = estimate, y = reorder(term_label, estimate), color = direction)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey45") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.15) +
  geom_point(size = 2.2) +
  scale_x_log10() +
  facet_wrap(~ isnv_set, scales = "free_y") +
  theme_minimal() +
  labs(x = "Rate ratio for iSNV count", y = NULL, color = "Direction")

p_lm_effects_sets <- lm_effects_by_set %>%
  mutate(term_label = case_when(
    term == "Ct" ~ "Ct",
    term == "logc" ~ "Average genome coverage (log10)",
    term == "mut_ivar" ~ "iVar mutation load",
    term == "age2older" ~ "Age: older (vs adult)",
    term == "age2young" ~ "Age: young (vs adult)",
    str_starts(term, "lineage_group") ~ str_replace(term, "lineage_group", "Variant: "),
    TRUE ~ term
  )) %>%
  ggplot(aes(x = estimate, y = reorder(term_label, estimate), color = direction)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey45") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.15) +
  geom_point(size = 2.2) +
  facet_wrap(~ isnv_set, scales = "free_y") +
  theme_minimal() +
  labs(x = "Linear-model effect on iSNV count", y = NULL, color = "Direction")

p_count_effects_sets / p_lm_effects_sets
```
## DN/DS analysis

we perform primary analysis of dn/ds using the primary data first and new CO_Infection calls
```{r dnds}

library(dndscv)
#sampleID chr      pos ref mut
names <- c(sampleID="barcode", chr="REGION", pos="POS",ref="REF",mut="ALT")
omicron=isnv_primary %>% filter(Ct<=25) %>% filter(lineage_group=="Omicron") %>% dplyr::select(barcode,REGION,POS,REF,ALT) %>% rename(all_of(names))

#library(dplyr)

#omicron <- omicron %>%
#  distinct(chr, pos, ref, mut, .keep_all = TRUE)

#we remove iSNVs happening at the same positiion within
other=isnv_primary %>% filter(Ct<=25) %>% filter(lineage_group=="Other") %>% dplyr::select(barcode,REGION,POS,REF,ALT) %>% rename(all_of(names)) #%>%  distinct(chr, pos, ref, mut, .keep_all = TRUE)
Lambda=isnv_primary %>% filter(Ct<=25) %>% filter(lineage_group=="Lambda") %>% dplyr::select(barcode,REGION,POS,REF,ALT) %>% rename(all_of(names)) #%>% distinct(chr, pos, ref, mut, .keep_all = TRUE)
Coi=isnv_primary %>% filter(Ct<=25) %>% filter(lineage_group=="Co-Infection") %>% dplyr::select(barcode,REGION,POS,REF,ALT) %>% rename(all_of(names)) #%>% distinct(chr, pos, ref, mut, .keep_all = TRUE)

# we run the dnds analisys 
dnds_omi = dndscv(omicron, refdb = "analysis/dndscv/db/RefCDS_MN908947.3.peptides.rda", max_muts_per_gene_per_sample = Inf, max_coding_muts_per_sample = Inf)
dnds_oth = dndscv(other, refdb = "analysis/dndscv/db/RefCDS_MN908947.3.peptides.rda", max_muts_per_gene_per_sample = Inf, max_coding_muts_per_sample = Inf)
dnds_lam = dndscv(Lambda, refdb = "analysis/dndscv/db/RefCDS_MN908947.3.peptides.rda", max_muts_per_gene_per_sample = Inf, max_coding_muts_per_sample = Inf)
dnds_coi = dndscv(Coi, refdb = "analysis//dndscv/db/RefCDS_MN908947.3.peptides.rda", max_muts_per_gene_per_sample = Inf, max_coding_muts_per_sample = Inf)

a=dnds_oth$sel_cv %>% dplyr::select(gene_name,wmis_cv) %>% mutate(var="Other")
b=dnds_lam$sel_cv %>% dplyr::select(gene_name,wmis_cv) %>% mutate(var="Lambda")
c=dnds_omi$sel_cv %>% dplyr::select(gene_name,wmis_cv) %>% mutate(var="Omicron")
d=dnds_coi$sel_cv %>% dplyr::select(gene_name,wmis_cv) %>% mutate(var="Co-Infection")

```


plot the results

```{r figdnds2}
# ---- Packages
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(patchwork)


# ---- 3) Aesthetic choices ----
variant_colors <- c(
  "Lambda" = "#8E7CC3",
  "Omicron" = "#B45F06",
  "Other" = "#38761D",
  "Co-Infection" = "#333333"
)

# ---- Global panel with CIs
global_dnds <- dplyr::bind_rows(
  dnds_oth$globaldnds[5,],
  dnds_lam$globaldnds[5,],
  dnds_omi$globaldnds[5,],
  dnds_coi$globaldnds[5,]
) %>%
  mutate(
    name = c("Other","Lambda","Omicron","Co-Infection"),
    name = factor(name, levels = c("Other","Lambda","Omicron","Co-Infection"))
  )

p_global <- ggplot(global_dnds, aes(x = name, y = mle, color = name)) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_pointrange(aes(ymin = cilow, ymax = cihigh), position = position_dodge(width = 0.2)) +
  scale_color_manual(values = variant_colors, guide = "none") + ylim(0,1.2) +
#  coord_flip() +
  labs(
    x = NULL,
    y = expression("Global " * omega ~ "(dN/dS, MLE)"),
    title = "Global dN/dS"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(size = 12),
    axis.title.y = element_text(margin = margin(r = 5)),
    axis.text.x = element_text(size = 10, angle = 20, hjust = 1)
  )


# ---- Combine (narrow global panel + wider gene panel)
#final_dn_ds_fig <- p_global + p_genes + plot_layout(widths = c(1, 3))
#final_dn_ds_fig
#pdf("FigX.pdf",height = 8,width = 16)
#final_dn_ds_fig
#dev.off()


# ---- 1) Combine & harmonize ----
# a, b, c, d must each have: gene_name, wmis_cv, var
df <- dplyr::bind_rows(a, b, c, d) %>%
  mutate(
    var = factor(var, levels = c("Other", "Lambda", "Omicron", "Co-Infection"))
  )

# ---- 2) Global gene order (same across all panels) ----
# Order by max dN/dS across variants (descending)
gene_levels <- df %>%
  group_by(gene_name) %>%
  summarize(max_omega = max(wmis_cv, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(max_omega)) %>%
  pull(gene_name)

df <- df %>%
  mutate(
    gene_name = factor(gene_name, levels = rev(gene_levels)),   # top = highest
    # Cap at 5 for plotting, flag those > 5
    omega_cap = pmin(wmis_cv, 3),
    capped    = wmis_cv > 3
  )



# Helper to make one panel; show_y controls gene labels
make_panel <- function(v, show_y = FALSE, show_x = FALSE) {
  ggplot(df %>% filter(var == v),
         aes(x = omega_cap, y = gene_name)) +
    geom_vline(xintercept = 1, linetype = "dashed", linewidth = 0.5) +
    geom_point(aes(shape = capped),
               size = 2.2,
               color = variant_colors[[as.character(v)]]) +
    scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17), guide = "none") +
    coord_cartesian(xlim = c(0, 3)) +
    labs(
      title = as.character(v),
      x = if (show_x) "dN/dS (missense, ω)" else NULL,
      y = if (show_y) "Gene" else NULL
    ) +
    theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(size = 12, hjust = 0),
      axis.text.y = if (show_y) element_text(size = 9) else element_blank(),
      axis.title.y = if (show_y) element_text() else element_blank(),
      panel.grid.minor = element_blank()
    )
}

# ---- 4) Build panels (labels only on the first) ----
p_other <- make_panel("Other",        show_y = TRUE)
p_lambda <- make_panel("Lambda",      show_y = FALSE)
p_omi   <- make_panel("Omicron",      show_y = FALSE, show_x = TRUE)
p_coi   <- make_panel("Co-Infection", show_y = FALSE)

# ---- 5) Combine with patchwork ----
gene_panels <- p_other | p_lambda | p_omi | p_coi
gene_panels

final_dn_ds_fig <- p_global + gene_panels + plot_layout(widths = c(1, 3))

pdf("Fig4_dnds.pdf",width = 10, height = 5)
final_dn_ds_fig
dev.off()

final_dn_ds_fig

var=df %>% select(gene_name,wmis_cv,var)%>% pivot_wider(names_from=var, values_from = wmis_cv)
global_dnds
var

```

Stadistical tests at gene-level between variants, 
we aim to anwser whether 
```{r statstest}
kruskal.test(as.numeric(wmis_cv) ~ var, data = df)
#pairwise test
FSA::dunnTest(wmis_cv ~ var, data=df, method="bh")

# accounting for gene differences
library(lme4)
m <- lmer(wmis_cv ~ var + (1 | gene_name), data=df)
anova(m)


```

