---
title: "Statistical Analysis COVID19"
author: "Pastor Jullian"
date: "2024-02-28"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Statistical analysis COVID-19

Analysis to identify regions of the SARS-CoV-2 genome that are different between groups.

#### Load libraries

```{r, message=FALSE}
library("ggpubr")
library('ggplot2')
library("dplyr")
```

### Data input
#### Numeric to factors:
From `mydata`, we want to modify columns into factors, with the exception of `value` column.
We also add a `window` column.

```{r}
mydata <- read.table('./heterogeneity_groups.csv', h=TRUE,sep=",", row.names = 1)
mydata <- mydata %>%
  mutate(across(where(is.character), as.factor))
mydata$window <- as.factor(1:30)
str(mydata)

```

### Data explaration and visualisation

```{r echo=FALSE}
# Plot
ggplot(mydata, aes(x = window, y = value, col = label))+
  geom_boxplot(width=0.7, position=position_dodge(width = 0.7))

# Density plot 
ggdensity(mydata$value, main = "Density plot",
          xlab = "Value")

# QQ plot
ggqqplot(mydata$value)

# Normality test
shapiro.test(mydata$value)

```
### Normalisation and lineal model

```{r eval=FALSE}

## Normalization
# 1. Log
mydata$logscale <- log(mydata$value)
shapiro.test(mydata$logscale)

# 2. Mean = 0  
mydata$scaled <- (mydata$value - mean(mydata$value)) / sd(mydata$value)
shapiro.test(mydata$scaled) 

## Linear model using:
# Raw values
lm1 <- 
  mydata %>% group_by(window) %>% 
  do(lm = summary(lm(value ~ label,data = .)))

# Log adjusted values
lm2 <- 
  mydata %>% group_by(window) %>% 
  do(lm = summary(lm(logscale ~ label,data = .)))

# Scaled values
lm3 <- 
  mydata %>% group_by(window) %>% 
  do(lm = summary(lm(scaled ~ label,data = .)))

```
### Adjustment for multiple comparison
1. First extract the p-values
2. Estimate FDR
3. Identify the significant comparison

```{r eval=FALSE}

# 1. Function to extract p-values
fun1 <- function(model){
pv <- list()
for (i in 1:dim(model)[1]) {
  pv[[i]] <- model$lm[[i]]$coefficients[,4]
  }
return(pv)
}

# 2. Function to estimate FDR
fun2 <- function(model){
    ad <- vector('list', length(model))
    for (i in 1:length(model)) {
    for (j in 1:3) {
      ad[[i]][j] <- p.adjust(model[[i]][j], "fdr",n = 3)
  }
  }
  return(ad)
}

# 3. Find the significant
fun3 <- function(model,cutoff){
bl <- list()
for (i in 1:length(model)) {
  bl[[i]] <- model[[i]] <= cutoff
  }
return(bl)
}

## FDR for:

# 1. Raw values
out1 <- fun1(lm1)
out2 <- fun2(out1)
out_raw <- fun3(out2,0.05)

# Significant windows: 29 and 30

# 2. Log values
out1 <- fun1(lm2)
out2 <- fun2(out1)
out_log <- fun3(out2,0.05)

# Significant windows: 23,27, 28, 29 and 30

# 3. Scale values
out1 <- fun1(lm3)
out2 <- fun2(out1)
out_scale <- fun3(out2,0.05)

# Significant windows: 29 and 30

```
### Result

From the results, we can see that groups from window 29 and 30 are significantly different, either using raw values or scaled values.


  