---
title: "Preprocessing of COVID samples"
output: rmarkdown::github_document
date: "2023-08-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

suppressMessages(library('tidyverse'))
suppressMessages(library('ggbeeswarm'))
suppressMessages(library('patchwork'))
suppressMessages(library('ggplot2'))
suppressMessages(library('reshape2'))
suppressMessages(library('ComplexHeatmap'))
suppressMessages(library('reticulate'))
use_python('/usr/local/bin/python3')
knitr::knit_engines$set(python = reticulate::eng_python)

```

## Processing results in python 

Cecret Nextflow pipeline was used to identify SARS-COV-2 variants in the samples. This pipeline contains multiple tools that allows us to detect variant types, mutations and heterogeneity in the samples. For this case we use the outcome from freyja and iVar.

In this step we take the tsv files that contains the results of the analysis of this tools and processing into tables to work with them further.

```{python, results='hide'}

import pandas as pd
import ast
import os 
#function to have the full path of each file in a folder
def listdir_fullpath(d):
    return [os.path.join(d, f) for f in os.listdir(d)]
  
#position variant from freya
freya_files = sorted(listdir_fullpath('join_flow_cells/freyja'))

with open('freya_results.txt', 'w') as f: #file to save results
    f.write('sampleId,th_filter,pass_filter')
    f.write('\n')
    for file in freya_files:
        
        if file.endswith('variants.tsv'): #open files where are the results of variants per position 
            data = pd.read_csv(file, sep='\t')
            #choosing to type of filters to select data
            th_filter = len(data[data["ALT_DP"]>400])
            pass_filter = len(data[data["PASS"] == True])
            sample_name = file.split('.')[0].split('/')[-1:][0][:-9] #sample name 
            f.write('{},{},{}'.format(sample_name,th_filter,pass_filter))
            f.write('\n')

#covid variant 
covid_variant = pd.read_csv('join_flow_cells/freyja/aggregated-freyja.tsv', sep='\t')
all_variants = {}
for index, row in covid_variant.iterrows():
    sample = 'sample_{}'.format(row['Unnamed: 0'].split('.')[0].split('_')[2])
    variants = ast.literal_eval(row['summarized'])
    all_variants[sample] = dict(variants)

all_variants = pd.DataFrame(all_variants)
all_variants.to_csv('freya_variants.tsv', sep='\t')

ivar_files = sorted(listdir_fullpath('join_flow_cells/ivar_variants'))
positions = {}

for file in ivar_files:   
    if file.endswith('variants.tsv'): #open files where are the results of variants per position 
        sample_name = 'sample_{}'.format(file.split('.')[0].split('/')[-1][-2:])
        data = pd.read_csv(file, sep='\t')
        count = {}
        for pos in data[['POS']].values:
            count[pos[0]] = 1
       
        positions[sample_name] = count

ivar_results = pd.DataFrame(positions).fillna(0).transpose()

ivar_results.to_csv('ivar_results.tsv', sep='\t', float_format='%.0f')

#
samtools_depth = sorted(listdir_fullpath('join_flow_cells/samtools_depth'))

samtools_data = {}
for file in samtools_depth:      
    if file.endswith('depth.txt'):
        file_name = file.split('.')[0].split('/')[-1:][0].split('_')[-1]
        sample_name = 'sample_{}'.format(file_name)[:-3]
        data = pd.read_csv(file, sep='\t', names=['ref', 'pos', 'depth'])
        #transforming data into a dictionary for easy transform into dataframe
        values = pd.Series(data.depth.values,index=data.pos).to_dict()
        samtools_data[sample_name] = values
        
pd.DataFrame(samtools_data).to_csv('samtools_depth.csv', sep=',')
```
## Joining data in R

After python codes generate tables from the iVar and Freyja output this are joined using R with the demographic information.

```{r loaddata, echo=FALSE}
#reading binary matrix of presence/absence of mutations detected by iVar 
ivar_results <- as.matrix(read.table('ivar_results.tsv', 
                           h=TRUE,sep="\t", check.names = FALSE, row.names = 1))
#deleting sample 71
ivar_results <- ivar_results[-71,]
#counting how many mutations by Ivar has every sample
n_mut_ivar <- data.frame(rowSums(ivar_results))
n_mut_ivar <- cbind(newColName = rownames(n_mut_ivar), n_mut_ivar)
rownames(n_mut_ivar) <- 1:nrow(n_mut_ivar)
colnames(n_mut_ivar) <- c('sample','num_mut')

#freya results
freya_results <-read.table('freya_results.txt', h=TRUE,sep=",")
freya_variants <-read.table('freya_variants.tsv', h=TRUE,sep="\t")
freya_variants <- freya_variants[,order(colnames(freya_variants))]
freya_variants <- melt(freya_variants)
#getting variant more prevalent
max <- freya_variants %>% group_by(variable) %>% top_n(1, value) 


#Demographic samples info
demographic_info <- read.table('/Users/bbernal/Documents/covid_work/demographic.csv', h=TRUE,sep=",")
demographic_info = demographic_info[c('sample', 'Comuna', 'Fecha.Análisis', 'Edad')]
demographic_info <- demographic_info %>% 
  mutate(ciudad = case_when(Comuna == 'RANCAGUA'  ~ 'Rancagua' ,
                            .default = 'Santiago'))

demographic_info <- demographic_info[order(match(demographic_info[,1],rownames(ivar_results))),]

demographic_info<-demographic_info[!grepl("71", 
                          demographic_info$sample),]

#adding freya results with sampleid names from ivar (normalizing sample names)
freya_results$sampleId <- n_mut_ivar$sample
#melt to plot 
freya_results <- melt(freya_results, id="sampleId")
freya_results <- filter(freya_results, variable == "pass_filter")

#joining all results
all_results <- cbind(freya_results, mut_ivar = n_mut_ivar$num_mut)
all_results$variant <- max$X
all_results$location <- demographic_info$ciudad
all_results$age <- demographic_info$Edad
all_results$analisis_data <- demographic_info$Fecha.Análisis
all_results <- all_results[,-2]
all_results <- all_results %>% rename("pass_freya" = "value")

head(all_results)

write.table(all_results, file='all_results.csv', sep=',', row.names=FALSE)

```

