---
title: "Analysis of COVID samples"
output: rmarkdown::github_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

suppressMessages(library('tidyverse'))
suppressMessages(library('ggbeeswarm'))
suppressMessages(library('patchwork'))
suppressMessages(library('ggplot2'))
suppressMessages(library('reshape2'))
suppressMessages(library('ComplexHeatmap'))
suppressMessages(library('rstatix'))
suppressMessages(library('ggpubr'))
suppressMessages(library('ggsignif'))
```

## Reading files

Reading results from iVar and freya from preprocessing output to generated figures.

```{r loaddata, echo=FALSE}
#Ivar results
#reading binary matrix of presence/absence of mutations detected by iVar 
ivar_results <- as.matrix(read.table('ivar_results.tsv', 
                           h=TRUE,sep="\t", check.names = FALSE, row.names = 1))
#deleting sample 71
ivar_results <- ivar_results[-71,]

all_results <- read.table(file='all_results.csv', sep=',', h=TRUE)
head(all_results)
```

## Results

### Mutation Heatmap

Using ComplexHeatmap library to plot if samples has 1 or any mutations detected by iVar, using a binary matrix. Variant a location is also included into the heatmap.

```{r, heatmapivar, echo=FALSE}
#colors setting 
variant <- setNames(RColorBrewer::brewer.pal(name = "Dark2", n = length(unique(all_results$variant))), unique(all_results$variant))
location = c("Rancagua" =  "royalblue", 
             "Santiago" = "brown3")
col.list <- list(variant = variant, location = location)

h_ann= rowAnnotation(location = all_results$location,
                        variant = all_results$variant,
                        col = col.list,
                        gp = gpar(fontsize = 3),
                        border = TRUE)
p1 <- Heatmap(ivar_results, 
               name = "Presence/Absence of mutations",
               rect_gp = gpar(col = "gray70", lwd = 0.1),
               row_names_gp = gpar(fontsize = 7, col ='gray9'),
               column_names_gp = gpar(fontsize = 4, col ='gray9'), # Text size for row names
               col = c("seagreen", "blue3"),
               row_km = 2,
               right_annotation = h_ann)

p1
```

### Number of variants vs heterogeneity 

Combining mutations detected by iVar and variations in to positions detected by Freya. Including variant information and location of samples.

```{r, mutations, echo=FALSE}
p2 <- ggplot(data=all_results, aes(x=mut_ivar, y=pass_freya, label=sampleId, shape=location, colour = variant)) +
  geom_point() +
  scale_color_manual('', values = col.list$variant) + 
  theme_minimal()

p2
```

## Heterogeneity between groups

In the previous there are three identifiable groups in the samples. Here is the analysis to check if in every variant posible in the genome and detected genes that are most mutated in the samples.

#### Groups:

Generate clusters trough kmeans with k = 3.
```{r, groups, echo=FALSE}
res.kmeans <- kmeans(all_results[,c("mut_ivar")], centers = 3, 
                     iter.max = 100, nstart = 50)
all_results$cluster <- factor(res.kmeans$cluster)
write.csv(all_results,file='all_results.csv', row.names=FALSE)


ggplot(data=all_results, aes(x=mut_ivar, y=pass_freya, label=sampleId, 
                             shape=location, colour = cluster)) +
  geom_point() + 
  theme(legend.text = element_text(size=0.2)) + 
  theme_minimal()

```
### Staticistic tests

First we check basics stats as mean and standard deviation
```{r, loadvariants, echo=FALSE}
pos_variant <- read.table('pos_variants.csv', h=TRUE,sep=",")
pos_variant <- melt(pos_variant)
pos_variant <-  left_join(pos_variant, all_results, by = c("variable" = "sampleId")) %>%
  select(contains(c('X', 'variable', 'value','cluster','variant','location','cluster')))

pos_variant %>%
  group_by(cluster) %>%
  get_summary_stats(value, type = "mean_sd")
```
First, it performs an anova to check the effects of cluster labeling into the variations in the genome position.
```{r, anova, echo=FALSE}
#primer approach
res.aov <- pos_variant %>% anova_test(value ~ cluster)
res.aov
```
Then to compare difference between groups it performs a t-test
```{r, t_test, echo=FALSE}

pwc <- pos_variant %>%
  pairwise_t_test(value ~ cluster, p.adjust.method = "bonferroni", detailed = TRUE)
pwc
pwc <- pwc %>% add_xy_position(x = "cluster")
```

Now we plot the data in a boxplot, showing anova result and t-test.
```{r, statsplot, echo=FALSE}
#test between groups
ggplot(pos_variant, aes(x=cluster, y=value, colour=factor(cluster))) + 
  geom_boxplot() + 
  stat_pvalue_manual(pwc, cluster = "p.adj", tip.length = 0, step.increase = 0.1) +
  labs(
    subtitle = get_test_label(res.aov, detailed = TRUE),
    caption = get_pwc_label(pwc)
  ) + theme_minimal()

```

## Samtools depth 

Checking depth of the samples and separate them by variant and location.

```{r, loadsamtools, echo=FALSE}
samtools_results <-read.table('/Users/bbernal/Documents/covid_work/samtools_depth.csv', h=TRUE,sep=",")
samtools_results <- melt(samtools_results, id="X")
colnames(samtools_results) <- c('pos', 'sampleId', 'depth')

samtools_results <- left_join(samtools_results,all_results[c('sampleId', 'location', 'variant')], by="sampleId")
samtools_results<-samtools_results[!grepl("71", samtools_results$sample),]

samtools_depth_range <- samtools_results %>%
  group_by(sampleId, grp = cut(pos, breaks=pretty(pos, n = 60)), location, variant) %>% 
  summarise(count = mean(depth))

```
### Samtools depth plot
```{r, depth, echo=FALSE}
p3 <- ggplot(samtools_depth_range , aes(x = grp, y=log(count),  
                                     colour = variant)) +
  geom_line(aes(group = sampleId), size = 0.3, alpha = 0.8)+
  geom_hline(yintercept = log(100),color="red2",
             linetype='dotted',size=0.5) +
  annotate("text", x = 45, y = log(100), label = "100X Coverage", vjust = -0.5, size=3) + 
  facet_grid(rows = vars(variant), col = vars(location)) +
  scale_color_manual('', values = col.list$variant) +
  theme_minimal() + theme(axis.text.x = element_text(angle = 90, hjust = 0.5, size = 5))

p3
```
