---
title: "Analysis of COVID samples"
output: rmarkdown::github_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

suppressMessages(library('tidyverse'))
suppressMessages(library('ggbeeswarm'))
suppressMessages(library('patchwork'))
suppressMessages(library('ggplot2'))
suppressMessages(library('reshape2'))
suppressMessages(library('ComplexHeatmap'))
suppressMessages(library('rstatix'))
suppressMessages(library('ggpubr'))
suppressMessages(library('ggsignif'))
suppressMessages(library('treeio'))
suppressMessages(library('ggtree'))
suppressMessages(library('muscle'))
suppressMessages(library('Biostrings'))

```

## Reading files

Reading results from iVar and freya from preprocessing output to generated figures.

```{r loaddata, echo=FALSE}
#Ivar results
#reading binary matrix of presence/absence of mutations detected by iVar 
ivar_results <- as.matrix(read.table('ivar_results.tsv', 
                           h=TRUE,sep="\t", check.names = FALSE, row.names = 1))
#deleting sample 71
ivar_results <- ivar_results[-71,]

all_results <- read.table(file='all_results.csv', sep=',', h=TRUE)
head(all_results)
```

## Results

### Mutation Heatmap

Using ComplexHeatmap library to plot if samples has 1 or any mutations detected by iVar, using a binary matrix. Variant and location is also included into the heatmap.

```{r, heatmapivar, echo=FALSE}
#colors setting 
variant <- setNames(RColorBrewer::brewer.pal(name = "Dark2", n = length(unique(all_results$variant))), unique(all_results$variant))
location = c("Rancagua" =  "royalblue", 
             "Santiago" = "brown3")
col.list <- list(variant = variant, location = location)

png(file="heatmap.png", width=22, height=20, units="cm", res=800)
h_ann= rowAnnotation(location = all_results$location,
                        variant = all_results$variant,
                        col = col.list,
                        gp = gpar(fontsize = 3),
                        border = TRUE)
p1 <- Heatmap(ivar_results, 
               name = "Presence/Absence of mutations",
               rect_gp = gpar(col = "gray70", lwd = 0.1),
               row_names_gp = gpar(fontsize = 7, col ='gray9'),
               column_names_gp = gpar(fontsize = 4, col ='gray9'), # Text size for row names
               col = c("seagreen", "blue3"),
               row_km = 2,
               right_annotation = h_ann)

p1

draw(p1)
dev.off()

```

### Number of variants vs heterogeneity 

Combining mutations detected by iVar and variations in to positions detected by Freya. Including variant information and location of samples.

```{r, mutations, echo=FALSE}
p2 <- ggplot(data=all_results, aes(x=mut_ivar, y=pass_freya, label=sampleId, shape=location, colour = variant)) +
  geom_point() +
  scale_color_manual('', values = col.list$variant) + 
  theme_minimal()

p2

ggsave("mutations.png", dpi=300, dev='png', height=8, width=14, units="cm")

```

## Heterogeneity between groups

In the previous there are three identifiable groups in the samples. Here is the analysis to check if in every variant posible in the genome and detected genes that are most mutated in the samples.

#### Groups:

Generate clusters trough kmeans with k = 3.
```{r, groups, echo=FALSE}
res.kmeans <- kmeans(all_results[,c("mut_ivar")], centers = 3, 
                     iter.max = 100, nstart = 50)
all_results$cluster <- factor(res.kmeans$cluster)
write.csv(all_results,file='all_results.csv', row.names=FALSE)


p3 <- ggplot(data=all_results, aes(x=mut_ivar, y=pass_freya, label=sampleId, 
                             shape=location, colour = cluster)) +
  geom_point() + 
  theme(legend.text = element_text(size=0.2)) + 
  theme_minimal()

p3

ggsave("cluster.png", dpi=300, dev='png', height=8, width=14, units="cm")
```
### Staticistic tests

First we check basics stats as mean and standard deviation
```{r, loadvariants, echo=TRUE}
pos_variant <- read.table('pos_variants.csv', h=TRUE,sep=",")
pos_variant <- melt(pos_variant)
pos_variant <-  left_join(pos_variant, all_results, by = c("variable" = "sampleId")) %>%
  select(contains(c('X', 'variable', 'value','cluster','variant','location','cluster')))

pos_variant %>%
  group_by(cluster) %>%
  get_summary_stats(value, type = "mean_sd")
```
First, it performs an anova to check the effects of cluster labeling into the variations in the genome position.
```{r, anova, echo=TRUE}
#primer approach
res.aov <- pos_variant %>% anova_test(value ~ cluster)
res.aov
```
Then to compare difference between groups it performs a t-test
```{r, t_test, echo=TRUE}

pwc <- pos_variant %>%
  pairwise_t_test(value ~ cluster, p.adjust.method = "bonferroni", detailed = TRUE)
pwc
pwc <- pwc %>% add_xy_position(x = "cluster")
```

Now we plot the data in a boxplot, showing anova result and t-test.
```{r, statsplot, echo=FALSE}
#test between groups
p5 <- ggplot(pos_variant, aes(x=cluster, y=value, colour=factor(cluster))) + 
  geom_boxplot() + 
  stat_pvalue_manual(pwc, cluster = "p.adj", tip.length = 0, step.increase = 0.1) +
  labs(
    subtitle = get_test_label(res.aov, detailed = TRUE),
    caption = get_pwc_label(pwc)
  ) + theme_minimal()

p5

ggsave("annova_test.png", dpi=300, dev='png', height=15, width=20, units="cm")
```

## Samtools depth 

Checking depth of the samples and separate them by variant and location.

```{r, loadsamtools, echo=TRUE}
samtools_results <-read.table('samtools_depth.csv', h=TRUE,sep=",")
samtools_results <- melt(samtools_results, id="X")
colnames(samtools_results) <- c('pos', 'sampleId', 'depth')

samtools_results <- left_join(samtools_results,all_results[c('sampleId', 'location', 'variant')], by="sampleId")
samtools_results<-samtools_results[!grepl("71", samtools_results$sample),]

samtools_results <- samtools_results %>% group_by(location, variant)

samtools_results <- samtools_results %>% group_by(location, variant) %>%
  mutate(med = mean(depth, na.rm = TRUE))

samtools_depth_range <- samtools_results %>%
  group_by(sampleId, grp = cut(pos, breaks=pretty(pos, n = 60), dig.lab = 5), location, variant, med) %>% 
  summarise(count = mean(depth, na.rm = TRUE))
```

### Samtools depth plot
```{r, depth, echo=FALSE}
p6 <- ggplot(samtools_depth_range , aes(x = grp, y=log(count),  
                                     colour = variant)) +
  geom_line(aes(group = sampleId), linewidth = 0.3, alpha = 0.8)+
  geom_hline(yintercept = log(100),color="red2",
             linetype='dotted',linewidth=0.5) + 
  annotate("text", x = 45, y = log(100), label = "100X Coverage", vjust = -0.5, size=3) +
  facet_grid(rows = vars(variant), col = vars(location)) +
  scale_color_manual('', values = col.list$variant) +
  geom_hline(aes(yintercept = log(med), group=interaction(location, variant)), 
             colour = 'darkblue', size =0.6, linetype='dashed') +
  geom_text(aes(y = log(med), x = 45, label = paste(round(med),"X Coverage"), 
                group=interaction(location, variant)),
                vjust = -0.6, size=2, color='gray40') + 
    theme_minimal() + 
  xlab("Position range (pb)") + ylab("Depth (log)") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 0.5, size = 5))

p6

ggsave("depth.png", dpi=300, dev='png', height=15, width=20, units="cm")
```

## Phylogenetic tree

Analysis of phylogeny of samples from results of iqtree and plotted with ggtree. We use as inspiration this figure https://yulab-smu.top/treedata-book/chapter13.html, and added the sequences distance to the tree. 

### Processing tree data

First, is necessary to transform tree data into a tiblle to add variant information from samples, after that it returns to treedata object.

```{r, phylotree, echo=TRUE}
my_tree <- read.newick('../Phylogenetics/iqtree2.treefile')
#transforming tree to add variant information
x <- as_tibble(my_tree)
#formatting sample name
x$sampleId <- paste0('sample_',(str_extract(x$label,"(?<=-)\\d+")))
#keeping name of reference
x <- x %>% 
  mutate(sampleId = ifelse(label == 'MN908947',label,sampleId))
#joining data
x_new <- treeio::full_join(x,all_results[c('sampleId', 'location', 'variant')], by="sampleId")
#transforming data into a tree object again!
x_new <- as.treedata(x_new)
```

#### First sight to phylogenetic tree

```{r, plottree, echo=TRUE}
#plot tree
p7 <- ggtree(x_new, color="grey40", ladderize = TRUE) +
  xlim_tree(0.00325) +
  geom_tiplab(aes(color=variant, label = sampleId), size=3.) +
  scale_color_manual('', values = variant) +
  guides(color = guide_legend(override.aes = list(size = 5, label = "\u25AA")))+
  theme_tree2(legend.position = "bottom")

p7
```

### Calculate distance

This is do it with MUSCLE package of R, the code is commented by the moment because it takes a lot of time to calculated, but we have alignment already saved. 
```{r, muscle, echo=FALSE}
### MUSCLE ANALYSIS - RUN ONE TIME ###
#extracting fasta files
#folder = "covid_genomics/Phylogenetics/consensus/"
#files = list.files(folder, pattern = ".fa")
#fa_seq = lapply(files,readDNAStringSet)
#fa_seq = do.call(c,fa_seq)
#tipseq_aln <- muscle::muscle(fa_seq) 
#saving file
#writeXStringSet(tipseq_aln, filepath = 'covid_genomics/Phylogenetics/samples_alignment.fasta', format = 'fasta')
```

#### Transforming data to be ready to plot

Is necessary to formatting data to plot and joined into the tree. 
```{r, musclefile, echo=TRUE}
#reading alignment file from MUSCLE
tipseq_aln <- readDNAStringSet("../Phylogenetics/samples_alignment.fasta")
tipseq_aln <- DNAStringSet(tipseq_aln)
#calculating distance between sample
tipseq_dist <- stringDist(tipseq_aln, method = "hamming")
## calculate the percentage of differences
tipseq_d <- as.matrix(tipseq_dist) / width(tipseq_aln[1]) * 100

## convert the matrix to a tidy data frame for facet_plot
dd <- as_tibble(tipseq_d)
dd$seq1 <- rownames(tipseq_d)
td <- gather(dd,seq2, dist, -seq1)
# with this we can save colors from tree
g <- p7$data$variant
names(g) <- p7$data$label
td$clade <- g[td$seq2] 

#this sample is deleted due to low coverage 
td<-td[!grepl("03-03", 
              c(td$seq1)),]
td<-td[!grepl("03-03", 
              c(td$seq2)),]

#adding a new column to set further order of apperance in the graph and to make that omicron variants keep together
td <- td %>%
  mutate( alpha = case_when(str_detect(clade, "Omicron") ~ "Omicron",
                            .default = clade))
```


### Phylogenetic tree and Distance sequence

```{r, seqdist, echo=TRUE}
#setting colors with the same palette 
variant <- setNames(RColorBrewer::brewer.pal(name = "Dark2", n = length(unique(all_results$variant))), unique(all_results$variant))
#sort to make alpha based on appearence 
td<- td[order(td$alpha), decreasing=TRUE]

#adding sequence distnace to phylo tree plot
p8 <- p7 + geom_facet(panel = "Sequence Distance", 
                     data = td, geom = geom_path,  size=0.35,
                     mapping=aes(x = dist, group = seq2, 
                                 color = clade, alpha = alpha),
                     show.legend = F)

#reshaping proportion of each plot
facet_labeller(p8, c(Tree = "phylogeny")) %>% facet_widths(c(Tree = 1.))

ggsave("PhylogenyIQTREE_SeqDist.png", dpi=300, dev='png', height=25, width=20, units="cm")
```