---
title: "Windows analysis"
output: rmarkdown::github_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressMessages(library('tidyverse'))
suppressMessages(library('patchwork'))
suppressMessages(library('ggplot2'))
suppressMessages(library('reshape2'))
suppressMessages(library('Gviz'))
suppressMessages(library('reticulate'))
py_install("pandas")

use_python('/usr/local/bin/python3')
knitr::knit_engines$set(python = reticulate::eng_python)
options(ucscChromosomeNames=FALSE)
```

# Analysis of windows

After the statistical test comparing the clusters defined and detecting significant windows in the genome, it calculated how many samples has variations in determinate positions in the genome and related to specifics genes, to do this this analysis is made in python and to incorporate in the plots the genome of covid it uses the library gviz.

## Getting number of samples with variations in the COVID genome

This code in python process all files from freyja to detect how many samples has variations/mutations in particular positions. Samples are separated in clusters that were previously detected with kmeans in 02_analysis and are saved in all_results.csv.

```{python}
import os 
import pandas as pd

def listdir_fullpath(d):
    return [os.path.join(d, f) for f in os.listdir(d)]
  
#reading data
all_results = pd.read_csv('all_results.csv')  

freya_files = sorted(listdir_fullpath('join_flow_cells/freyja'))

#separating clusters 
cluster1 = list(all_results[(all_results['cluster'] == 1)]['sampleId'])
cluster2 = list(all_results[(all_results['cluster'] == 2)]['sampleId'])
cluster3 = list(all_results[(all_results['cluster'] == 3)]['sampleId'])

cluster_lists = {'cluster1': cluster1,
'cluster2': cluster2,
'cluster3': cluster3}

samples_per_position = {}
for cluster_name,cluster_list in cluster_lists.items():
    cluster = {}
    for sample in cluster_list:
        for file in freya_files:
            if file.endswith('variants.tsv'):
                file_name = file.split('.')[0].split('/')[-1:][0][:-9].split('_')[-1]
                #checking files per clusters 
                sample_name = 'sample_{}'.format(file_name)[:-3]
                if sample == sample_name:

                    data = pd.read_csv(file, sep='\t')
                    pass_filter = data[data["PASS"] == True]
                    #saving positions detected per sample file
                    for pos in pass_filter['POS'].values:
                        if pos not in cluster:
                            cluster[pos] = []
                        cluster[pos].append(sample_name)

    samples_per_position[cluster_name] = cluster

#checking positions and saving samples if has variations in specific positions
count_per_pos = {}
for cluster,list_count in samples_per_position.items():
    cluster_count = {}
    for pos,samples in list_count.items():
        new_values = set(samples)
        cluster_count[pos] = len(new_values)
    
    count_per_pos[cluster] = cluster_count
    
count_per_pos_table= pd.DataFrame.from_dict(count_per_pos)

pd.DataFrame.from_dict(count_per_pos_table).to_csv('windows_cluster.csv', sep=',')
```

## Gviz 

Uploading data of positions.
```{r windows_data, echo=FALSE}
all_results <-read.table('all_results.csv', h=TRUE,sep=",")
windows <-read.table('windows_cluster.csv', h=TRUE,sep=",")
#sorting windows to be in numerical order
windows <- windows[order(windows$X),]
#replacing na values with zeros
windows[is.na(windows)] <- 0
head(windows)

#getting coordinates to be accepted in gviz
coords <- as.numeric(windows$X)
coords <- append(coords, 23000)
coords <- sort(coords)

#separating cluster and normalizing by number of samples in each cluster
n_samples <- table(all_results$cluster)
cluster1 <- windows$cluster1/n_samples[1]
cluster2 <- windows$cluster2/n_samples[2]
cluster3 <- windows$cluster3/n_samples[3]

```

## Setting variable to gviz
```{r gviz_settings, echo=TRUE}
dtrack1 <- DataTrack(data = cluster1, start = coords[-length(coords)], 
                     end = coords[-1], chromosome = 'NC_045512v2', genome = 'wuhCor1', 
                     groups = factor("cluster 1", 
                              levels = c("cluster 1", "cluster 2", "cluster 3")),
                     name = "#samples",
                     legend = TRUE)

dtrack2 <- DataTrack(data = cluster2, start = coords[-length(coords)], 
                     end = coords[-1], chromosome = 'NC_045512v2', genome = 'wuhCor1', 
                     groups = factor("cluster 2", 
                                     levels = c("cluster 1", "cluster 2", "cluster 3")), 
                     name = "#samples",
                     legend = TRUE)

dtrack3 <- DataTrack(data = cluster3, start = coords[-length(coords)], 
                     end = coords[-1], chromosome = 'NC_045512v2', genome = 'wuhCor1', 
                     groups = factor("cluster 3", 
                                     levels = c("cluster 1", "cluster 2", "cluster 3")), 
                     name = "#samples",
                     legend = TRUE)

#size of dots and width of lines 
displayPars(dtrack1) <- list(lwd = 3, cex=0.7, alpha=0.9)
displayPars(dtrack2) <- list(lwd = 3, cex=0.7, alpha=0.9)
displayPars(dtrack3) <- list(lwd = 3, cex=0.7, alpha=0.9)

#genes to be ploted
gtrack <- GenomeAxisTrack(range = IRanges(start = c(19620, 21560,26245, 26523, 28274), 
                                          end = c(20650, 25384, 26472, 27191, 29533),
                                          names = c('Endonucleasa', 'Spike', "E", "M", "N")))

#join dtrack data
ot <- OverlayTrack(trackList=list(dtrack1, dtrack2, dtrack3), name = "# samples")
#covid genome
idxTrack <- IdeogramTrack(genome="wuhCor1")
#box to highlight windows o interest
ht1 <- HighlightTrack(trackList=list(idxTrack, gtrack, ot), 
                      start = c(22000, 26000, 27000), end = c(23000, 27000, 28000), 
                      chromosome = "NC_045512v2")

print(dtrack1)
print(idxTrack)
```

### All genome with higlight in specific window
```{r plot1, echo=TRUE}
plotTracks(list(ht1),
           from = 0, to = 29900, 
           type =  c("p", "smooth"), window=500, showId = TRUE)
```

### Specific window 
```{r plot2, echo=TRUE}
plotTracks(list(ht1),
           start = c(22000, 26000, 27000), end = c(23000, 27000, 28000), 
           type =  c("p", "smooth"), window=50, showId = TRUE)

```