---
title: "analysis_covid"
output: rmarkdown::github_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::current_input(dir = TRUE)

suppressMessages(library('tidyverse'))
suppressMessages(library('openxlsx'))
suppressMessages(library('ggbeeswarm'))
suppressMessages(library('patchwork'))
suppressMessages(library('ggplot2'))
suppressMessages(library('reshape2'))
suppressMessages(library('ComplexHeatmap'))

```

## Reading files

Reading results from iVar and freya from join_flow_cell pipeline and including demographic information of samples.

```{r loaddata, echo=FALSE}
#Ivar results
ivar_files <- list.files(path = "analysis/join_flow_cells/ivar_variants/", pattern = "\\.tsv$", recursive = T, full.names = TRUE)

sample <- ivar_files[0:96]
ivar_results <- list()
for(i in sample){
  sample = paste('sample_', 
                 strsplit(strsplit(strsplit(i, '/')[[1]][10], '_')[[1]][3],  '.', fixed = TRUE)[[1]][1], 
                 sep = '')
  variant <- substring(sample, 1, nchar(sample)-3)
  df <- read.table(i, h=TRUE,sep="\t")
  ivar_results[[variant]] <- df$POS}

#generation of binary matrix 
mut_matrix <- t(splitstackshape:::charMat(listOfValues = ivar_results, fill = 0L))

colnames(mut_matrix) <- names(ivar_results)
mut_matrix <- mut_matrix[order(rowSums(mut_matrix),decreasing=T),] #sort by position with more mutations
mut_matrix <- t(mut_matrix) #transpose
#deleting sample 71
mut_matrix <- mut_matrix[-71,]

#counting how many mutations by Ivar has every sample
n_mut_ivar <- data.frame(rowSums(mut_matrix))
n_mut_ivar <- cbind(newColName = rownames(n_mut_ivar), n_mut_ivar)
rownames(n_mut_ivar) <- 1:nrow(n_mut_ivar)
colnames(n_mut_ivar) <- c('sample','num_mut')
```

```{r loaddemographic, echo=FALSE}
#Demographic samples info
demographic_info <- read.table('analysis/demographic.csv', h=TRUE,sep=",")
demographic_info = demographic_info[c('sample', 'Comuna', 'Fecha.Análisis', 'Edad')]
demographic_info <- demographic_info %>% 
  mutate(ciudad = case_when(Comuna == 'RANCAGUA'  ~ 'Rancagua' ,
                            .default = 'Santiago'))

demographic_info <- demographic_info[order(match(demographic_info[,1],rownames(mut_matrix))),]
```

```{r loadfreya, echo=FALSE}
#freya results
freya_results <-read.table('/analysis/join_flow_cells/freya_results.txt', h=TRUE, sep=",")
freya_variants <-read.table('/analysis/join_flow_cells/freya_variants.tsv', h=TRUE,sep="\t")
freya_variants <- freya_variants[,order(colnames(freya_variants))]
freya_variants <- melt(freya_variants)
#getting variant more prevalent
max <- freya_variants %>% group_by(variable) %>% top_n(1, value) 

#deleting sample 71 
n_mut_ivar<-n_mut_ivar[!grepl("71", 
                              n_mut_ivar$sample),]
demographic_info<-demographic_info[!grepl("71", 
                          demographic_info$sample),]
#adding freya results with sampleid names from ivar (normalizing sample names)
freya_results$sampleId <- n_mut_ivar$sample
#melt to plot 
freya_results <- melt(freya_results, id="sampleId")
freya_results <- filter(freya_results, variable == "pass_filter")
```

```{r joinresults, echo=FALSE}
#joining all results
all_results <- cbind(freya_results, mut_ivar = n_mut_ivar$num_mut)
all_results$variant <- max$X
all_results$location <- demographic_info$ciudad
all_results$age <- demographic_info$Edad
all_results$analisis_data <- demographic_info$Fecha.Análisis

```

## Results

### Mutation Heatmap

Using ComplexHeatmap library to plot if samples has 1 or any mutations detected by iVar, using a binary matrix. Variant a location is also included into the heatmap.

```{r, heatmapivar, echo=FALSE}
#colors setting 
variant <- setNames(RColorBrewer::brewer.pal(name = "Accent", n = length(unique(all_results$variant))), unique(all_results$variant))
location = c("Rancagua" =  "royalblue", 
             "Santiago" = "brown3")
col.list <- list(variant = variant, location = location)

h_ann= rowAnnotation(location = all_results$location,
                        variant = all_results$variant,
                        col = col.list,
                        gp = gpar(fontsize = 3),
                        border = TRUE)
p1 <- Heatmap(mut_matrix, 
               name = "iVar mutation",
               rect_gp = gpar(col = "gray70", lwd = 0.1),
               row_names_gp = gpar(fontsize = 7, col ='gray9'),
               column_names_gp = gpar(fontsize = 4, col ='gray9'), # Text size for row names
               col = c("seagreen", "blue3"),
               row_km = 2,
               right_annotation = h_ann)

p1
```

### Number of variants vs heterogeneity 

Combining mutations detected by iVar and variations in to positions detected by Freya. Including variant information and location of samples.

```{r, mutations, echo=FALSE}
p2 <- ggplot(data=all_results, aes(x=mut_ivar, y=value, label=sampleId, shape=location,       colour = variant)) +
  geom_point() +
  scale_color_manual('', values = col.list$variant) + 
  theme_minimal()

p2
```

### Heterogeneity between groups

In the previous there are three identifiable groups in the samples. Here is the analysis to check if in every variant posible in the genome and detected genes that are most mutated in the samples.

#### Groups:

```{r, groups, echo=FALSE}

p2 + geom_vline(xintercept = 21,color="red2",
                       linetype='dotted',size=0.5) +
            geom_vline(xintercept = 40,color="red2",
                        linetype='dotted',size=0.5)

```

##Samtools depth 

Checking depth of the samples and separate them by variant and location 

```{r, loadsamtools, echo=FALSE}
samtools_files <- list.files(path = "/analysis/join_flow_cells/samtools_depth/", pattern = "\\.txt$", recursive = T, full.names = TRUE)
sample <- samtools_files[0:96]

depth <- list()
for(i in sample){
  #transform name of sample to sample_NN
  sample = paste('sample_',strsplit(strsplit(strsplit(i, '/')[[1]][10], '_')[[1]][3], '.', fixed = TRUE)[[1]][1], sep = '')
  variant <- substring(sample, 1, nchar(sample)-3)
  df <- read.table(i, h=FALSE,sep="\t")
  colnames(df) <- c('ref', 'pos', 'depth')
  depth[[variant]] <- df}

samtools_results <- lapply(depth, as.data.frame) %>% 
  bind_rows(.id = "sampleId")
samtools_results <- left_join(samtools_results,all_results[c('sampleId', 'location', 'variant')], by="sampleId")
samtools_results <- samtools_results[,-2]
samtools_results<-samtools_results[!grepl("71", 
                                          samtools_results$sample),]

samtools_depth_range <- samtools_results %>%
  group_by(sampleId, grp = cut(pos, breaks=pretty(pos, n = 60)), location, variant) %>% 
  summarise(count = mean(depth))

```
### Samtools depth plot
```{r, depth, echo=FALSE}
p3 <- ggplot(samtools_depth_range , aes(x = grp, y=log(count),  
                                     colour = variant)) +
  geom_line(aes(group = sampleId), size = 0.3, alpha = 0.8)+
  geom_hline(yintercept = log(100),color="red2",
             linetype='dotted',size=0.5) +
  annotate("text", x = 45, y = log(100), label = "100X Coverage", vjust = -0.5, size=3) + 
  facet_grid(rows = vars(variant), col = vars(location)) +
  scale_color_manual('', values = col.list$variant) +
  theme_minimal() + theme(axis.text.x = element_text(angle = 90, hjust = 0.5, size = 5))

p3
```
