---
title: "analysis_covid"
output: rmarkdown::github_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::current_input(dir = TRUE)

suppressMessages(library('tidyverse'))
suppressMessages(library('openxlsx'))
suppressMessages(library('ggbeeswarm'))
suppressMessages(library('patchwork'))
suppressMessages(library('ggplot2'))
suppressMessages(library('reshape2'))
suppressMessages(library('ComplexHeatmap'))
suppressMessages(library('here'))


```

## Reading files

Reading results from iVar and freya from join_flow_cell pipeline and including demographic information of samples.

```{r loaddata, echo=FALSE}
#Ivar results

ivar_files <- list.files(path = "/join_flow_cells/ivar_variants/", pattern = "\\.tsv$")

sample <- ivar_files[0:96]
ivar_results <- list()
for(i in sample){
  sample = paste('sample_', 
                 strsplit(strsplit(i, '_')[[1]][3],  '.', fixed = TRUE)[[1]][1], 
                 sep = '')
  variant <- substring(sample, 1, nchar(sample)-3)
  df <- read.table(i, h=TRUE,sep="\t")
  ivar_results[[variant]] <- df$POS}

#generation of binary matrix 
mut_matrix <- t(splitstackshape:::charMat(listOfValues = ivar_results, fill = 0L))
colnames(mut_matrix) <- names(ivar_results)
mut_matrix <- mut_matrix[order(rowSums(mut_matrix),decreasing=T),] #sort by position with more mutations
mut_matrix <- t(mut_matrix) #transpose
#deleting sample 71
mut_matrix <- mut_matrix[-71,]

#counting how many mutations by Ivar has every sample
n_mut_ivar <- data.frame(rowSums(mut_matrix))
n_mut_ivar <- cbind(newColName = rownames(n_mut_ivar), n_mut_ivar)
rownames(n_mut_ivar) <- 1:nrow(n_mut_ivar)
colnames(n_mut_ivar) <- c('sample','num_mut')

#Demographic samples info
demographic_info <- read.table('/demographic.csv', h=TRUE,sep=",")
demographic_info = demographic_info[c('sample', 'Comuna', 'Fecha.Análisis', 'Edad')]
demographic_info <- demographic_info %>% 
  mutate(ciudad = case_when(Comuna == 'RANCAGUA'  ~ 'Rancagua' ,
                            .default = 'Santiago'))

demographic_info <- demographic_info[order(match(demographic_info[,1],rownames(mut_matrix))),]

#freya results
freya_results <-read.table('/join_flow_cell/freya_results.txt', h=TRUE,sep=",")
freya_variants <-read.table('/join_flow_cell/freya_variants.tsv', h=TRUE,sep="\t")
freya_variants <- freya_variants[,order(colnames(freya_variants))]
freya_variants <- melt(freya_variants)
#getting variant more prevalent
max <- freya_variants %>% group_by(variable) %>% top_n(1, value) 

#deleting sample 71 
n_mut_ivar<-n_mut_ivar[!grepl("71", 
                              n_mut_ivar$sample),]
demographic_info<-demographic_info[!grepl("71", 
                          demographic_info$sample),]
#adding freya results with sampleid names from ivar (normalizing sample names)
freya_results$sampleId <- n_mut_ivar$sample
#melt to plot 
freya_results <- melt(freya_results, id="sampleId")
freya_results <- filter(freya_results, variable == "pass_filter")

#joining all results
all_results <- cbind(freya_results, mut_ivar = n_mut_ivar$num_mut)
all_results$variant <- max$X
all_results$location <- demographic_info$ciudad
all_results$age <- demographic_info$Edad
all_results$analisis_data <- demographic_info$Fecha.Análisis

```

### Mutation Heatmap

Using ComplexHeatmap library to plot if samples has 1 or any mutations detected by iVar, using a binary matrix. Variant a location is also included into the heatmap.

```{r, heatmapivar, echo=FALSE}

h_ann= rowAnnotation(location = all_results$location,
                        variant = all_results$variant,
                        col = list(location = c("Rancagua" =  "royalblue", 
                                            "Santiago" = "brown3"), 
                                   variant = c("Other" = "#DF536B",
                                               "Gamma" = "#28E2E5",
                                               "Lambda" = "#2297E6",
                                               "Epsilon" = "gray62",
                                               "Alpha" = "#CD0BBC",              
                                               "BA.2* [Omicron (BA.2.X)]" =  "#F5C710",
                                               "Omicron" = "#61D04F" )),
                        gp = gpar(fontsize = 3),
                        border = TRUE)
p1 <- Heatmap(mut_matrix, 
               name = "iVar mutation",
               rect_gp = gpar(col = "gray70", lwd = 0.1),
               row_names_gp = gpar(fontsize = 7, col ='gray9'),
               column_names_gp = gpar(fontsize = 4, col ='gray9'), # Text size for row names
               col = c("seagreen", "blue3"),
               row_km = 2,
               right_annotation = h_ann)

p1
```

### Number of variants vs heterogeneity 

Combining mutations detected by iVar and variations in to positions detected by Freya. Including variant information and location of samples.

```{r, mutations, echo=FALSE}
p2 <- ggplot(data=all_results, aes(x=mut_ivar, y=value, label=sampleId, 
                             colour = variant, shape=location)) + 
  geom_point() +
  theme_minimal()

p2
```