---
title: "Covid Figures"
author: "Alex Di Genova"
output: rmarkdown::github_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Figure 1
Demographic of sequenced COVID samples.
## Chilean map

Map of Chile highlithing the Rancagua y Santiago regions
```{r chimap, echo=FALSE}
# Cargar los paquetes necesarios
library(ggplot2)
library(rnaturalearth)

# Obtener los datos geográficos de Chile
chile_map <- ne_countries(country = "chile", returnclass = "sf")

# Coordenadas aproximadas de Santiago y Rancagua
locations <- data.frame(
  city = c("Santiago", "Rancagua"),
  lon = c(-70.6483, -70.7398),
  lat = c(-33.4569, -34.1708)
)

# Crear el mapa utilizando ggplot2 y resaltar Santiago y Rancagua
p1=ggplot(data = chile_map) +
  geom_sf() +
  geom_point(data = locations, aes(x = lon, y = lat, color = city), size = 3) +
  scale_color_manual(values = c("Santiago" = "blue", "Rancagua" = "red")) +
  ggtitle("Chile") +
  theme_minimal() +
  theme(text = element_text(size = 12), legend.position = "none",  axis.text.x=element_blank()) +
  labs(color = "Ciudad")
p1
```

## Demographic of samples

```{r demog}
library(tidyverse)
df=read.table("../demographics/all_samples.txt", h=T)
dates_formatted <- as.Date(df$Date, format="%m/%d/%Y")
years <- format(dates_formatted, "%Y")
df$year=years
df=df %>% mutate(Ct=as.integer(Ct))
df$all="all"
```
we make the plot

```{r demogp}
library(tidyverse)
library(ggbeeswarm)
count_data <- df %>%
  group_by(year) %>%
  summarise(count = n())

p2=ggplot(df, aes(x = year, y = Age, )) +
     #geom_boxplot(aes(group = year), color="brown", alpha = 0.5) + # Add boxplot
  geom_quasirandom(aes(color = Region, shape = Sex),width = 0.25, height = 0, size=3) +
  # geom_beeswarm(aes(color = Region, shape = Sex),width = 0.25, height = 0, size=3) +
   #geom_jitter(, width = 0.2, height = 0) +
  scale_color_manual(values = c("Metropolitana" = "blue", "O'Higgins" = "red")) +
  labs(title = "Samples",
       x = "Year",
       y = "Age") +
    theme(text = element_text(size = 12)) +
  theme_minimal()
p2

# Ct plots 2.1
#ggplot(df, aes(x=all,shape=Sex,y=Ct,color=Region,group=all)) + geom_quasirandom() + theme_minimal()
# Ct plots 2.2
#ggplot(df, aes(x=Age,shape=Sex,y=Ct,color=Region,group=all)) + geom_point() + theme_minimal()

```


## GISAD

```{r data, echo=FALSE}
#reading data
all_results <-read.table('all_results.csv', h=TRUE,sep=",")
metadata = read.table('GISAID_all_chile.tsv', h=TRUE,sep="\t", check.names = FALSE)

#formatting dates
metadata$Date <-ym(strtrim(metadata$`Collection date`, 7))
variants <- metadata[,c("Lineage","Date")]

#extracting dates of sampling data
samples_dates <- ym(strtrim(all_results$analisis_data, 7))
samples_dates <- na.omit(samples_dates)
samples_dates <- unique(samples_dates)

#generating table for bar plot
hist_variants <- variants %>%
  group_by(Date) %>% 
  count(Lineage)

#formating lineage info to variant names, some links for manual search of VOI and VOC
#https://www.cdc.gov/coronavirus/2019-ncov/variants/variant-classifications.html
#https://gisaid.org/hcov19-variants/

hist_variants <- hist_variants %>%
  mutate(
    name = case_when(
      str_detect(Lineage, "P.2") ~ "Zeta",
      str_detect(Lineage, "B.1.617.1") ~ "Kappa",
      str_detect(Lineage, "B.1.526") ~ "Iota",
      str_detect(Lineage, "B.1.525") ~ "Eta",
      str_detect(Lineage, "B.1.427|B.1.429") ~ "Epsilon",
      str_detect(Lineage, "B.1.1.529|BA.2.86|CH.1.1|BA.2.74|^BA.\\d|XBB.1.\\d") ~ "Omicron",
      str_detect(Lineage, "B.1.617|B.1.617.2|^AY.\\d") ~ "Delta",
      str_detect(Lineage, "B.1.1.7|^Q.\\d") ~ "Alpha",
      str_detect(Lineage, "P.1|^P.1.\\d") ~ "Gamma",
      str_detect(Lineage, "B.1.351|B.1.351.2|B.1.351.3") ~ "Beta",
      str_detect(Lineage, "C.37|C.37.1") ~ "Lambda",
      str_detect(Lineage, "B.1.621|B.1.621.1") ~ "Mu",
      .default = "Other"
    )
  )

#calculating percentage of variants in each month
variant_percentage <- hist_variants %>% 
  group_by(Date) %>%
  count(name) %>%
  mutate(percent = n/sum(n))
```

## Plotting number of samples and percentage of variants in Chile.

## Number of samples

Here we got the number of samples in Chile from GISAID platform, we label this with their corresponding name of VOI or VOC. With blue arrows we point out where our samples are located on the timeline. Around years 2021 and 2022 there is huge increase in the sequencing of COVID samples in Chile.


```{r num_samples, echo=FALSE}
# Number of samples 
p3 <- ggplot(hist_variants) + 
  geom_bar(aes(x=Date, y=n, fill=name), stat="identity") +
  annotate("segment", x=samples_dates, xend=samples_dates,
    y=1000, yend=2, color="red",
    arrow=arrow(length=unit(0.4,"cm"))) +
  scale_x_date(date_labels="%Y",date_breaks  ="1 year") +
  theme_minimal()+
  labs(title = "GISAD Chilean Genomes",
       x = "Date",
       y = "# Covid Genomes",
       fill = "Variants") +
  theme(text = element_text(size = 12))
  #theme(text = element_text(size = 12), legend.position = "bottom")
p3
```

We compose all plots to generate figure 1

```{r cmposer}

library(patchwork)

patchwork=p1|(p2/p3)  
patchwork + plot_annotation(tag_levels = 'A')
pdf(file="Fig1.pdf",height=7,width=9)
patchwork + plot_annotation(tag_levels = 'A')
dev.off()

```

# Supplementary Figure 1

Details about the sequenced reads, including number of reads per sample, read length and read quality, adaptor sequence content ect.

```{r supf1}
library(ggbeeswarm)
library(tidyverse)
mq_jfc = read.table('../figure_qc/join_flow_cells/multiqc_samtools_stats.txt', header = TRUE)
mq_jfc$barcode=as.integer(str_extract(mq_jfc$Sample,"(?<=-)\\d+"))
mq_jfc=left_join(mq_jfc,df,by=c("barcode"="barcode"))
mq_jfc = mq_jfc %>% dplyr::select(Sample,
                           raw_total_sequences,
                           error_rate,average_quality,
                           insert_size_average,
                           reads_mapped_percent,
                           reads_properly_paired_percent,
                           reads_unmapped_percent,
                           Ct)
mq_jfc = mq_jfc %>% mutate(raw_total_sequences=raw_total_sequences/1000000)
order_fct=c("Ct",
            "raw_total_sequences",
            "average_quality",
            "insert_size_average",
            "reads_mapped_percent",
            "reads_unmapped_percent",
            "reads_properly_paired_percent",
            "error_rate")
order_fct_r=c("PCR Ct",
              "Raw total reads (M)", 
              "Average read quality (Q-score)",
              "Insert size average (bp)" ,
              "Reads mapped (%)",
              "Reads unmapped (%)",
              "Read properly paired (%)",
              "Error rate (%)")
colors <- setNames(RColorBrewer::brewer.pal(name = "Dark2", n =8),order_fct_r)
#pivot longer
mq_jfc=mq_jfc %>% pivot_longer(-Sample)

mq_jfc$rename=setNames(order_fct_r, order_fct)[mq_jfc$name]

sf1=mq_jfc %>% 
  mutate(name = fct_relevel(rename, order_fct_r)) %>%
  ggplot(aes(y=rename,x=value,color=rename,group=name)) + 
  geom_violin()+
  geom_quasirandom(orientation="y") + 
  scale_color_manual(values=colors)+
  stat_summary(fun = median, geom = "vline", aes(xintercept = ..x..), linetype = 2, linewidth= 0.8, color = "black") +
  ylab("")  +
  xlab("")+
   theme_minimal()+ facet_wrap(.~name,scales="free",ncol=2,dir="v") +  
  theme(legend.position = "none",
        axis.text.y = element_blank(), 
        strip.text.x = element_text(
        size = 14,
        color = "black", 
        face = "bold", 
        hjust = 0)
        )

sf1
pdf("SF1.pdf")
sf1
dev.off()
```

We create a table to give the information in main text
```{r, statsreads}
mq_jfc %>% group_by(rename) %>% summarise(total=sum(value),median=median(value),mean=mean(value),min=min(value),max=max(value),sd=sd(value))
```

# Figure 2

Philogenetics, variants detected and coverage.

## Phylogenetic tree
```{r readdata,echo=FALSE}
#BiocManager::install("treeio")
#BiocManager::install("ggtree")

suppressMessages(library('treeio'))
suppressMessages(library('ggtree'))
suppressMessages(library(phytools))

all_results <- read.table(file='all_results.csv', sep=',', h=TRUE)

my_tree <- read.newick('../Phylogenetics/iqtree2.treefile')
rerooted_tree <- reroot(my_tree, node.number = 1)
#transforming tree to add variant information
x <- as_tibble(rerooted_tree)
#formatting sample name
x$sampleId <- paste0('sample_',(str_extract(x$label,"(?<=-)\\d+")))
#keeping name of reference
x <- x %>% 
  mutate(sampleId = ifelse(label == 'MN908947',label,sampleId))
#joining data
x_new <- treeio::full_join(x,all_results[c('sampleId', 'location', 'variant')], by="sampleId")
#transforming data into a tree object again!
x_new <- as.treedata(x_new)
#we rename the tips
a=x_new@phylo$tip.label
a=paste0('sample_',(str_extract(a,"(?<=-)\\d+")))
a[1]="MN908947"
x_new@phylo$tip.label=a
variant <- setNames(RColorBrewer::brewer.pal(name = "Dark2", n = length(unique(c(all_results$variant,"Co-Infection")))),
                    unique(c(all_results$variant,"Co-Infection")))

ivar_results <- as.matrix(read.table('ivar_results.tsv', 
                           h=TRUE,sep="\t", check.names = FALSE, row.names = 1))
#deleting sample 71
ivar_results <- ivar_results[-71,]

pos_order=as.character(sort(as.numeric(colnames(ivar_results))))
#sort by coordinate
ivar_sor=ivar_results[,pos_order]
ivdt=as.data.frame(ivar_sor)
ivdt$name=rownames(ivdt)
ivdt_longer=ivdt%>%pivot_longer(!name, names_to = "pos", values_to = "snp") %>% mutate(pos=as.numeric(pos)) %>% filter(snp==1)
a=rowSums(ivar_results)
ivar_c=data.frame(name=names(a),value=a)

```

We plot the tree including variants type, iVAR snp distribution and number of SNPs


```{r plott}

library(ggtree)

# Customize further as needed
p4 <- ggtree(x_new, color="grey40", ladderize = TRUE) +
  #xlim_tree(0.00325) +
  geom_tiplab(aes(color=variant, label = sampleId), size=2.) +
  scale_color_manual('', values = variant) +
  scale_fill_manual('', values = variant) +
  guides(color = guide_legend(override.aes = list(size = 5, label = "\u25AA")))+
  theme_tree2(legend.position = "bottom")

# add SNP data by position in the tree
p5= p4 + geom_facet(panel = "SNPs coordinates", data = ivdt_longer, geom = geom_point, 
               mapping=aes(x = pos,group=sampleId, color=variant,fill=variant), shape = '|') +
             geom_facet(panel = "# SNPs", data = ivar_c, geom = geom_col, 
                aes(x = value, color = variant, fill=variant), orientation = 'y', width = .6)
p5

#facet_widths(p8, widths = c(2, 2, 1))
pdf(file="Fig2.pdf",height=7,width=9)
facet_widths(p5, widths = c(2, 2, 1))
dev.off()
```




# Figure 3

Genome coverage

## Samtools depth 

Checking depth of the samples and separate them by variant and location.

```{r, loadsamtools, echo=TRUE}
suppressMessages(library('reshape2'))
samtools_results <-read.table('samtools_depth.csv', h=TRUE,sep=",")
samtools_results <- melt(samtools_results, id="X")
colnames(samtools_results) <- c('pos', 'sampleId', 'depth')

samtools_results <- left_join(samtools_results,all_results[c('sampleId', 'location', 'variant')], by="sampleId")
samtools_results<-samtools_results[!grepl("71", samtools_results$sample),]

samtools_results <- samtools_results %>% group_by(variant)

samtools_results <- samtools_results %>% group_by(variant) %>%
  mutate(med = mean(depth, na.rm = TRUE))

samtools_depth_range <- samtools_results %>%
  group_by(sampleId, grp = cut(pos, breaks=pretty(pos, n = 60), dig.lab = 5),  variant, med) %>% 
  summarise(count = mean(depth, na.rm = TRUE))
```

### Samtools depth plot
```{r, depth, echo=FALSE}



p6 <- ggplot(samtools_depth_range , aes(x = grp, y=log(count),  
                                     colour = variant)) +
  geom_line(aes(group = sampleId), linewidth = 0.3, alpha = 0.8)+
  #geom_hline(yintercept = log(100),color="red2",
   #          linetype='dotted',linewidth=0.5) + 
  #annotate("text", x = 45, y = log(100), label = "100X Coverage", vjust = -0.5, size=2) +
  ylim(8,14)+
  scale_x_discrete(labels=as.character(seq(0,30000,500)))+
  facet_grid(rows = vars(variant),scales="fixed") +
  scale_color_manual('', values = variant) +
  geom_hline(aes(yintercept = log(med), group=variant), 
             colour = 'darkblue', size =0.6, linetype='dashed') +
  geom_text(aes(y = log(med), x = 45, label = paste(round(med),"X Coverage"), 
                group=variant),
                vjust = -0.6, size=4, color='gray40') + 
    theme_minimal() + 
  xlab("Position range (pb)") + ylab("Depth (log)") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 0.5, size = 5), legend.position = "bottom")

p6

pdf(file="Fig3.pdf",height=7,width=7)
p6
dev.off()
```



# Figure 4

Heterogeneity analysis

## SNVs

```{r loadsnv}
library(tidyverse)
csv_files <- list.files(pattern="*.variants.tsv",path = "../SNVs/")
l_data <- list()
for (file in csv_files) {
  data <- read.table(paste0("../SNVs/",file),h=T)
  #data <- data %>% filter(ALT_FREQ >=0.95)
  data$Filename <- file
  l_data[[file]] <- data
}
SNV_ivar <- bind_rows(l_data)
SNV_ivar$barcode=as.integer(str_extract(SNV_ivar$Filename,"(?<=-)\\d+"))
```

## iSNVs

loading a cleaning iSNV data, our filters include ALT_FREQ in [0.05 to 0.95], ALT_quality > 30, ALT_DP >100 as well as virus load <= 28, and no co-infection.

```{r loadisnvs}
csv_files <- list.files(pattern="*_variants.tsv",path = "../iSNVs/")
filtered_data <- list()
for (file in csv_files) {
  data <- read.table(paste0("../iSNVs/",file),h=T)
  dataf=data %>% filter(ALT_DP >=100) %>% filter(PASS=="TRUE")%>% filter(ALT_FREQ >= 0.05 & ALT_FREQ<0.95 & ALT_QUAL > 30)
  dataf$Filename <- file
  filtered_data[[file]] <- dataf
}
#dim(iSNVs)
iSNVs <- bind_rows(filtered_data)
iSNVs$barcode=as.integer(str_extract(iSNVs$Filename,"(?<=-)\\d+"))
#we remove all variants calls by ivar in the same sample/position
iSNVs=iSNVs %>% filter(!paste0(barcode,POS,ALT) %in% paste0(SNV_ivar$barcode,SNV_ivar$POS,SNV_ivar$ALT))

iSNVs_count = iSNVs %>%group_by(Filename) %>% count()
iSNVs_count$barcode=as.integer(str_extract(iSNVs_count$Filename,"(?<=-)\\d+"))
iSNV_data=left_join(iSNVs_count,df) %>% mutate(age2=if_else(Age < 30,"young",if_else(Age >=30 & Age<50,"adult","older")))

b=all_results %>% mutate(barcode=as.integer(str_extract(sampleId,"(?<=_)\\d+")))  %>% dplyr::select(barcode,mut_ivar,variant)
iSNV_data=left_join(iSNV_data,b)
iSNV_data=iSNV_data %>% mutate(variant=if_else(variant=="BA.2* [Omicron (BA.2.X)]","Omicron",variant))

# Ct <= 28
iSNV_data=iSNV_data%>% filter(Ct<=28)
# filtro de Co-infection muestras con mayor de 3% de por mas de otra variante
ff=read.table("aggregated-freyja-parse.tsv",h=T,sep="\t")
fff=ff%>%filter(percentage*100 > 3) %>%dplyr::select(file) %>% group_by(file) %>% count() %>% filter(n>1)

library(dplyr)

# Example input (your data frame)
# ff <- read.table("your_table.tsv", header = TRUE, sep = "\t")

# Filter and summarize
co_infect_table <- ff %>%
  filter(percentage > 0.03) %>%                  # Keep relevant variants
  group_by(file) %>%
  summarise(
    n_variants = n_distinct(variant),            # Number of variants detected
    variants = paste(unique(variant), collapse = ", ")  # Variant names
  ) %>%
  mutate(
    co_infection = ifelse(n_variants > 1, "Yes", "No")
  ) %>%
  arrange(desc(n_variants)) %>% filter(n_variants > 1)

# View result
print(co_infect_table)



fff$barcode=as.integer(str_extract(fff$file,"(?<=-)\\d+"))
iSNV_coi=iSNV_data %>% filter(barcode %in% fff$barcode)
iSNV_data=iSNV_data %>% filter(!barcode %in% fff$barcode)
iSNV_coi$variant="Co-Infection"
iSNV_data=rbind(iSNV_data,iSNV_coi)
# load coverage data
gcov=read.table("resume_samtools_coverage.txt",h=T)
gcov=gcov %>% mutate(barcode=as.integer(str_extract(sname,"(?<=-)\\d+"))) %>% dplyr::select(barcode,meandepth,coverage) %>%mutate(logc=log10(meandepth))
iSNV_data=left_join(iSNV_data,gcov)
iSNV_coi=left_join(iSNV_coi,gcov)

# we filter the iSNVs position relative to the samples filtered in the previos steps
iSNVs=iSNVs %>% filter(barcode %in% iSNV_data$barcode)
iSNVs=left_join(iSNVs,iSNV_data %>% dplyr::select(barcode,variant,Sex,meandepth), by=c("barcode"="barcode"))


# improve co-infection panel plot
#pfrya=ff %>% mutate(variant=if_else(str_detect(variant,"Omicron"),"Omicron",variant)) %>% filter(percentage*100 > 3)%>% ggplot(aes(x=fct_reorder2(file,desc(percentage),variant),y=percentage,fill=variant)) + geom_col() + theme_minimal() + labs(x="Samples", y="% variant") + theme( axis.text.x=element_blank())
```

### Comparison iSNV vs SNV

SNVs increase as function of time but iSNVs do not increase with time. Indicating other forces

```{r cmpsnvs}
library(ggpmisc)

MutSNV=SNV_ivar %>% count(barcode) %>% inner_join(.,df)
p1=MutSNV %>% filter(Ct<=28) %>% ggplot(aes(x=Sex,y=n,color = Region)) + geom_quasirandom() + 
  facet_wrap(~year)+
   scale_color_manual(values = c("Metropolitana" = "blue", "O'Higgins" = "red")) + 
  labs(x="Sexo",y="SNV",color = "Region",title="SNVs per year")+
  theme_light() +theme(legend.position = "none") 

p2=iSNV_data %>% filter(Ct<=28) %>% ggplot(aes(x=Sex,y=n,color = Region)) + geom_quasirandom() + 
  facet_wrap(~year)+
   scale_color_manual(values = c("Metropolitana" = "blue", "O'Higgins" = "red")) + 
  labs(x="Sexo",y="iSNV",color = "Region",title="iSNVs per year")+
  theme_light() +theme(legend.position = "none") 
#p1/p2


### 

p3=MutSNV %>% filter(Ct<=28) %>%mutate(Date = as.Date(Date,format = "%m/%d/%Y"))%>% ggplot(aes(x=Date,y=n)) + geom_point(aes(color=Region),size=2,alpha = 0.9) + geom_smooth(method = "lm", se = TRUE) + 
   stat_poly_eq(
    aes(label = paste( after_stat(eq.label),after_stat(rr.label), sep = "~~~")),
    formula = y ~ x,
    parse = TRUE
  ) +
  labs(title = "SNVs vs Date") + 
  scale_x_date(date_breaks = "4 months", date_labels = "%b/%Y") +
  labs(x="Date",y="SNV")+
  scale_color_manual(values = c("Metropolitana" = "blue", "O'Higgins" = "red")) + 
  theme_light() +theme(legend.position = "none") 

p4=iSNV_data %>% filter(Ct<=28) %>%mutate(Date = as.Date(Date,format = "%m/%d/%Y"))%>% ggplot(aes(x=Date,y=n)) + geom_point(aes(color=Region),size=2,alpha = 0.9) + geom_smooth(method = "lm", se = TRUE) + 
   stat_poly_eq(
    aes(label = paste( after_stat(eq.label),after_stat(rr.label), sep = "~~~")),
    formula = y ~ x,
    parse = TRUE
  ) +
  labs(title = "iSNVs vs Date") + 
  scale_x_date(date_breaks = "4 months", date_labels = "%b/%Y") +
  labs(x="Date",y="iSNV")+
  scale_color_manual(values = c("Metropolitana" = "blue", "O'Higgins" = "red")) + 
  theme_light() +theme(legend.position = "bottom") 

(p1/p2)|(p3/p4) 

pdf(file="SNV_iSNV_date.pdf",width = 8, height =6)
(p1/p2)|(p3/p4) 
dev.off()
```

```{r}
MutSNV=MutSNV %>% filter(Ct<=28) %>%mutate(Date = as.Date(Date,format = "%m/%d/%Y"))
summary(lm(n~Date,data=MutSNV))

iSNV_data=iSNV_data %>% filter(Ct<=28) %>%mutate(Date = as.Date(Date,format = "%m/%d/%Y"))
summary(lm(n~Date,data=iSNV_data))
```

### Regression model

```{r}
iSNVlm=lm(iSNV_data$n~iSNV_data$age2+iSNV_data$Sex
          +iSNV_data$mut_ivar+iSNV_data$variant+iSNV_data$Ct+iSNV_data$logc)
summary(iSNVlm)
```


```{r lmmedian_test}
# Packages
library(ggpubr)
library(broom)
library(dplyr)
library(ggplot2)
library(glue)
library(patchwork)

#=============================
# A) Linear model forest plot
#=============================
fit <- iSNVlm

coef_df <- tidy(fit, conf.int = TRUE) %>%
  dplyr::filter(term != "(Intercept)") %>%
  dplyr::mutate(
    term_label = dplyr::case_when(
      term == "iSNV_data$age2older"            ~ "Age: older (vs ref)",
      term == "iSNV_data$age2young"            ~ "Age: young (vs ref)",
      term == "iSNV_data$SexM"                 ~ "Sex: male (vs ref)",
      term == "iSNV_data$Ct"                   ~ "PCR Ct value",
      term == "iSNV_data$logc"                 ~ "Average genome coverage (log)",
      term == "iSNV_data$mut_ivar"             ~ "iVar mutation load",
      term == "iSNV_data$variantCo-Infection"  ~ "Variant: co-infection",
      term == "iSNV_data$variantEpsilon"       ~ "Variant: Epsilon",
      term == "iSNV_data$variantGamma"         ~ "Variant: Gamma",
      term == "iSNV_data$variantLambda"        ~ "Variant: Lambda",
      term == "iSNV_data$variantOmicron"       ~ "Variant: Omicron",
      term == "iSNV_data$variantOther"         ~ "Variant: Other",
      TRUE ~ term
    ),
    sig = dplyr::case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      p.value < 0.1   ~ "·",
      TRUE            ~ ""
    ),
    term_label = reorder(term_label, estimate)
  )

gl <- glance(fit)
subtitle_txt <- glue(
  "Adj. R² = {round(gl$adj.r.squared, 3)};  F({gl$df.null - gl$df.residual}, {gl$df.residual}) = ",
  "{round(gl$statistic, 2)}, p = {signif(gl$p.value, 3)}"
)

p_model <- ggplot(coef_df, aes(x = term_label, y = estimate)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2, linewidth = 0.6) +
  geom_point(size = 2.3) +
  geom_text(aes(label = sig, y = conf.high), vjust = -0.6, size = 4) +
  coord_flip() +
  labs(
    title = "A",
    subtitle = "Predictors of iSNV count (β, 95% CI)",
    x = NULL, y = "Effect on iSNV count",
    #caption = "Stars: *** p<0.001, ** p<0.01, * p<0.05, · p<0.10"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(size = 16, hjust = 0, face = "plain"),
    plot.subtitle = element_text(color = "grey20", size = 11),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 10),
    axis.title.y = element_text(size = 11)
  )

#=============================
# B) Sex comparison (Wilcoxon)
#=============================
sex_comparisons <- list(c("F","M"))

p_sex <- ggboxplot(
  iSNV_data, x = "Sex", y = "n",
  color = "Sex", palette = "jco",
  add = "jitter", #size = 1.8
) +
  stat_compare_means(
    comparisons = sex_comparisons,
    method = "wilcox.test", exact = FALSE,
    label = "p.format"
  ) +
  labs(title = "B", x = "Sex", y = NULL) +
  ylim(30, 150) +
  theme_pubr(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 16, hjust = -0.1, face = "plain")
  )

#=============================
# C) Variant comparison (K–W + pairs)
#=============================
mvars <- c("Lambda", "Omicron", "Other", "Co-Infection")
iSNV_data_tmp <- iSNV_data %>% dplyr::filter(variant %in% mvars)

pairwise_comparisons <- list(
 # c("Lambda", "Omicron"),
  c("Other", "Lambda"),
  c("Omicron", "Other"),
  c("Co-Infection", "Omicron"),
  c("Co-Infection", "Other"),
  c("Co-Infection", "Lambda")
)

variant_colors <- c(
  "Lambda" = "#8E7CC3",
  "Omicron" = "#B45F06",
  "Other" = "#38761D",
  "Co-Infection" = "#333333"
)

p_variant <- ggboxplot(
  iSNV_data_tmp, x = "variant", y = "n",
  color = "variant", palette = variant_colors,
  add = "jitter", #size = 1.8, width = 0.6
) +
  stat_compare_means(comparisons = pairwise_comparisons, exact = FALSE, label = "p.format") +
  stat_compare_means(label.y = 145, method = "kruskal.test", label = "p.format") +
  labs(title = "C", x = "SARS-CoV-2 variant", y = NULL) +
  ylim(30, 150) +
  theme_pubr(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 10, angle = 20, hjust = 1),
    axis.title.x = element_text(size = 11, margin = margin(t = 10)),
    plot.title = element_text(size = 16, hjust = -0.1, face = "plain")
  )

#=============================
# Shared y-axis label for B+C
#=============================
y_shared <- ggplot() +
  annotate("text", x = 0, y = 0,
           label = "Number of iSNVs",
           angle = 90, size = 4) +
  theme_void() +
  theme(plot.margin = margin(r = 2))

# Stack B over C (remove their y labels); prepend the shared y-label column
right_block <- (p_sex / p_variant) + plot_layout(heights = c(1, 1))
right_block_with_ylabel <- y_shared | right_block
right_block_with_ylabel <- right_block_with_ylabel + plot_layout(widths = c(0.06, 1))

#=============================
# Final 2-column figure: A | (B/C with shared y)
#=============================
final_fig <- p_model | right_block_with_ylabel
final_fig <- final_fig + plot_layout(widths = c(1.05, 1.35))

pdf("Fig4_new.pdf",width = 8, height = 6)
final_fig
dev.off()
final_fig




```

Mutation types distribution and count 

```{r aleles}
library(ggridges)
tmp=iSNVs %>% dplyr::select("REGION","POS","REF", "ALT","ALT_FREQ","variant") %>% filter(variant %in% c("Co-Infection","Omicron","Lambda","Other")) %>% mutate(allele=paste0(REF,"->",ALT))
tmp2=tmp
tmp2$variant="all"
tmp=rbind(tmp,tmp2)

tmp$allele=factor(tmp$allele,levels=names(sort(table(tmp$allele))))


  

p1=tmp  %>% ggplot(aes(y=allele,x=ALT_FREQ,fill=allele,color=allele))  +  geom_density_ridges() + theme_ridges(center_axis_labels=TRUE) + facet_wrap(.~variant,ncol=1) + theme(legend.position = "none") + labs(x="Frequency", y="Mutation types")

p2=tmp %>% ggplot(aes(y=allele,x=ALT_FREQ,fill=allele,color=allele))  +  geom_col() +  facet_wrap(.~variant,ncol=1) + theme_ridges(center_axis_labels=TRUE) + theme(legend.position = "none",axis.text.y = element_blank(),axis.title.y = element_blank()) + labs(x="Count") 

pdf("Fig5_new.pdf",width = 7, height = 11)
p1+p2 + plot_layout(widths = c(3,1))
dev.off()
p1+p2 + plot_layout(widths = c(3,1))
```

Mutation impact and Dn/Ds per gene in samples

### running dndscv on covid data

```{r dndscovid}
#library(devtools); install_github("im3sanger/dndscv")

library(dndscv)
#sampleID chr      pos ref mut
names <- c(sampleID="barcode", chr="REGION", pos="POS",ref="REF",mut="ALT")
omicron=iSNVs %>% filter(variant=="Omicron") %>% dplyr::select(barcode,REGION,POS,REF,ALT) %>% rename(all_of(names))

#library(dplyr)

omicron <- omicron %>%
  distinct(chr, pos, ref, mut, .keep_all = TRUE)

#we remove iSNVs happening at the same positiion within
other=iSNVs %>% filter(variant=="Other") %>% dplyr::select(barcode,REGION,POS,REF,ALT) %>% rename(all_of(names)) %>%  distinct(chr, pos, ref, mut, .keep_all = TRUE)
Lambda=iSNVs %>% filter(variant=="Lambda") %>% dplyr::select(barcode,REGION,POS,REF,ALT) %>% rename(all_of(names)) %>% distinct(chr, pos, ref, mut, .keep_all = TRUE)
Coi=iSNVs %>% filter(variant=="Co-Infection") %>% dplyr::select(barcode,REGION,POS,REF,ALT) %>% rename(all_of(names)) %>% distinct(chr, pos, ref, mut, .keep_all = TRUE)

# we run the dnds analisys 
dnds_omi = dndscv(omicron, refdb = "./dndscv/db/RefCDS_MN908947.3.peptides.rda", max_muts_per_gene_per_sample = Inf, max_coding_muts_per_sample = Inf)
dnds_oth = dndscv(other, refdb = "./dndscv/db/RefCDS_MN908947.3.peptides.rda", max_muts_per_gene_per_sample = Inf, max_coding_muts_per_sample = Inf)
dnds_lam = dndscv(Lambda, refdb = "./dndscv/db/RefCDS_MN908947.3.peptides.rda", max_muts_per_gene_per_sample = Inf, max_coding_muts_per_sample = Inf)
dnds_coi = dndscv(Coi, refdb = "./dndscv/db/RefCDS_MN908947.3.peptides.rda", max_muts_per_gene_per_sample = Inf, max_coding_muts_per_sample = Inf)

a=dnds_oth$sel_cv %>% dplyr::select(gene_name,wmis_cv) %>% mutate(var="Other")
b=dnds_lam$sel_cv %>% dplyr::select(gene_name,wmis_cv) %>% mutate(var="Lambda")
c=dnds_omi$sel_cv %>% dplyr::select(gene_name,wmis_cv) %>% mutate(var="Omicron")
d=dnds_coi$sel_cv %>% dplyr::select(gene_name,wmis_cv) %>% mutate(var="Co-Infection")

```
```{r figdnds2}
# ---- Packages
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(patchwork)


# ---- 3) Aesthetic choices ----
variant_colors <- c(
  "Lambda" = "#8E7CC3",
  "Omicron" = "#B45F06",
  "Other" = "#38761D",
  "Co-Infection" = "#333333"
)

# ---- Global panel with CIs
global_dnds <- dplyr::bind_rows(
  dnds_oth$globaldnds[5,],
  dnds_lam$globaldnds[5,],
  dnds_omi$globaldnds[5,],
  dnds_coi$globaldnds[5,]
) %>%
  mutate(
    name = c("Other","Lambda","Omicron","Co-Infection"),
    name = factor(name, levels = c("Other","Lambda","Omicron","Co-Infection"))
  )

p_global <- ggplot(global_dnds, aes(x = name, y = mle, color = name)) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_pointrange(aes(ymin = cilow, ymax = cihigh), position = position_dodge(width = 0.2)) +
  scale_color_manual(values = variant_colors, guide = "none") + ylim(0,1.2) +
#  coord_flip() +
  labs(
    x = NULL,
    y = expression("Global " * omega ~ "(dN/dS, MLE)"),
    title = "Global dN/dS"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(size = 12),
    axis.title.y = element_text(margin = margin(r = 5)),
    axis.text.x = element_text(size = 10, angle = 20, hjust = 1)
  )


# ---- Combine (narrow global panel + wider gene panel)
#final_dn_ds_fig <- p_global + p_genes + plot_layout(widths = c(1, 3))
#final_dn_ds_fig
#pdf("FigX.pdf",height = 8,width = 16)
#final_dn_ds_fig
#dev.off()


# ---- 1) Combine & harmonize ----
# a, b, c, d must each have: gene_name, wmis_cv, var
df <- dplyr::bind_rows(a, b, c, d) %>%
  mutate(
    var = factor(var, levels = c("Other", "Lambda", "Omicron", "Co-Infection"))
  )

# ---- 2) Global gene order (same across all panels) ----
# Order by max dN/dS across variants (descending)
gene_levels <- df %>%
  group_by(gene_name) %>%
  summarize(max_omega = max(wmis_cv, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(max_omega)) %>%
  pull(gene_name)

df <- df %>%
  mutate(
    gene_name = factor(gene_name, levels = rev(gene_levels)),   # top = highest
    # Cap at 5 for plotting, flag those > 5
    omega_cap = pmin(wmis_cv, 3),
    capped    = wmis_cv > 3
  )



# Helper to make one panel; show_y controls gene labels
make_panel <- function(v, show_y = FALSE, show_x = FALSE) {
  ggplot(df %>% filter(var == v),
         aes(x = omega_cap, y = gene_name)) +
    geom_vline(xintercept = 1, linetype = "dashed", linewidth = 0.5) +
    geom_point(aes(shape = capped),
               size = 2.2,
               color = variant_colors[[as.character(v)]]) +
    scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17), guide = "none") +
    coord_cartesian(xlim = c(0, 3)) +
    labs(
      title = as.character(v),
      x = if (show_x) "dN/dS (missense, ω)" else NULL,
      y = if (show_y) "Gene" else NULL
    ) +
    theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(size = 12, hjust = 0),
      axis.text.y = if (show_y) element_text(size = 9) else element_blank(),
      axis.title.y = if (show_y) element_text() else element_blank(),
      panel.grid.minor = element_blank()
    )
}

# ---- 4) Build panels (labels only on the first) ----
p_other <- make_panel("Other",        show_y = TRUE)
p_lambda <- make_panel("Lambda",      show_y = FALSE)
p_omi   <- make_panel("Omicron",      show_y = FALSE, show_x = TRUE)
p_coi   <- make_panel("Co-Infection", show_y = FALSE)

# ---- 5) Combine with patchwork ----
gene_panels <- p_other | p_lambda | p_omi | p_coi
gene_panels

final_dn_ds_fig <- p_global + gene_panels + plot_layout(widths = c(1, 3))

pdf("Fig4_dnds.pdf",width = 10, height = 5)
final_dn_ds_fig
dev.off()

final_dn_ds_fig

var=df %>% select(gene_name,wmis_cv,var)%>% pivot_wider(names_from=var, values_from = wmis_cv)
global_dnds
var

```

Stadistical tests at gene-level between variants, 
we aim to anwser whether 
```{r statstest}
kruskal.test(as.numeric(wmis_cv) ~ var, data = df)
#pairwise test
FSA::dunnTest(wmis_cv ~ var, data=df, method="bh")

# accounting for gene differences
library(lme4)
m <- lmer(wmis_cv ~ var + (1 | gene_name), data=df)
anova(m)


```

There is no global difference in dN/dS (missense burden) between variants at the gene level.

A p-value of 0.94 indicates the distributions of dN/dS across Other, Lambda, Omicron, and Co-infection are essentially indistinguishable when considering all genes together.


Across the SARS-CoV-2 genome, variants share similar overall selective constraints at the gene level.

This is consistent with the known biology:

Most SARS-CoV-2 genes evolve under strong purifying selection (dN/dS < 1).

Adaptive changes occur in specific loci (e.g., Spike, ORF6, ORF10) rather than globally.

Omicron, despite having many mutations, does not show a uniform shift in ω across all genes.

Co-infections (which often involve Omicron + another lineage) do not show additive or synergistic increases in selective pressure.

These results align with studies indicating that SARS-CoV-2 diversification involves localized episodic selection, not widespread genome-wide shifts in evolutionary pressure.

