---
title: "Covid Figures"
author: "Alex Di Genova"
output: rmarkdown::github_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Figure 1
Demographic of sequenced COVID samples.
## Chilean map

Map of Chile hihglithing the Rancagua y Santiago regions
```{r chimap, echo=FALSE}
# Cargar los paquetes necesarios
library(ggplot2)
library(rnaturalearth)

# Obtener los datos geográficos de Chile
chile_map <- ne_countries(country = "chile", returnclass = "sf")

# Coordenadas aproximadas de Santiago y Rancagua
locations <- data.frame(
  city = c("Santiago", "Rancagua"),
  lon = c(-70.6483, -70.7398),
  lat = c(-33.4569, -34.1708)
)

# Crear el mapa utilizando ggplot2 y resaltar Santiago y Rancagua
p1=ggplot(data = chile_map) +
  geom_sf() +
  geom_point(data = locations, aes(x = lon, y = lat, color = city), size = 3) +
  scale_color_manual(values = c("Santiago" = "blue", "Rancagua" = "red")) +
  ggtitle("Chile") +
  theme_minimal() +
  theme(text = element_text(size = 12), legend.position = "none",  axis.text.x=element_blank()) +
  labs(color = "Ciudad")
p1
```

## Demographic of samples

```{r demog}
df=read.table("../demographics/all_samples.txt", h=T)
dates_formatted <- as.Date(df$Date, format="%m/%d/%Y")
years <- format(dates_formatted, "%Y")
df$year=years
```
we make the plot

```{r demogp}
library(tidyverse)
library(ggbeeswarm)
count_data <- df %>%
  group_by(year) %>%
  summarise(count = n())

p2=ggplot(df, aes(x = year, y = Age, )) +
     #geom_boxplot(aes(group = year), color="brown", alpha = 0.5) + # Add boxplot
  geom_quasirandom(aes(color = Region, shape = Sex),width = 0.25, height = 0, size=3) +
  # geom_beeswarm(aes(color = Region, shape = Sex),width = 0.25, height = 0, size=3) +
   #geom_jitter(, width = 0.2, height = 0) +
  scale_color_manual(values = c("Metropolitana" = "blue", "O'Higgins" = "red")) +
  labs(title = "Samples",
       x = "Year",
       y = "Age") +
    theme(text = element_text(size = 12)) +
  theme_minimal()
p2
```


## GISAD

```{r data, echo=FALSE}
#reading data
all_results <-read.table('all_results.csv', h=TRUE,sep=",")
metadata = read.table('GISAID_all_chile.tsv', h=TRUE,sep="\t", check.names = FALSE)

#formatting dates
metadata$Date <-ym(strtrim(metadata$`Collection date`, 7))
variants <- metadata[,c("Lineage","Date")]

#extracting dates of sampling data
samples_dates <- ym(strtrim(all_results$analisis_data, 7))
samples_dates <- na.omit(samples_dates)
samples_dates <- unique(samples_dates)

#generating table for bar plot
hist_variants <- variants %>%
  group_by(Date) %>% 
  count(Lineage)

#formating lineage info to variant names, some links for manual search of VOI and VOC
#https://www.cdc.gov/coronavirus/2019-ncov/variants/variant-classifications.html
#https://gisaid.org/hcov19-variants/

hist_variants <- hist_variants %>%
  mutate(
    name = case_when(
      str_detect(Lineage, "P.2") ~ "Zeta",
      str_detect(Lineage, "B.1.617.1") ~ "Kappa",
      str_detect(Lineage, "B.1.526") ~ "Iota",
      str_detect(Lineage, "B.1.525") ~ "Eta",
      str_detect(Lineage, "B.1.427|B.1.429") ~ "Epsilon",
      str_detect(Lineage, "B.1.1.529|BA.2.86|CH.1.1|BA.2.74|^BA.\\d|XBB.1.\\d") ~ "Omicron",
      str_detect(Lineage, "B.1.617|B.1.617.2|^AY.\\d") ~ "Delta",
      str_detect(Lineage, "B.1.1.7|^Q.\\d") ~ "Alpha",
      str_detect(Lineage, "P.1|^P.1.\\d") ~ "Gamma",
      str_detect(Lineage, "B.1.351|B.1.351.2|B.1.351.3") ~ "Beta",
      str_detect(Lineage, "C.37|C.37.1") ~ "Lambda",
      str_detect(Lineage, "B.1.621|B.1.621.1") ~ "Mu",
      .default = "Other"
    )
  )

#calculating percentage of variants in each month
variant_percentage <- hist_variants %>% 
  group_by(Date) %>%
  count(name) %>%
  mutate(percent = n/sum(n))
```

## Plotting number of samples and percentage of variants in Chile.

## Number of samples

Here we got the number of samples in Chile from GISAID platform, we label this with their corresponding name of VOI or VOC. With blue arrows we point out where our samples are located on the timeline. Around years 2021 and 2022 there is huge increase in the sequencing of COVID samples in Chile.


```{r num_samples, echo=FALSE}
# Number of samples 
p3 <- ggplot(hist_variants) + 
  geom_bar(aes(x=Date, y=n, fill=name), stat="identity") +
  annotate("segment", x=samples_dates, xend=samples_dates,
    y=1000, yend=2, color="red",
    arrow=arrow(length=unit(0.4,"cm"))) +
  scale_x_date(date_labels="%Y",date_breaks  ="1 year") +
  theme_minimal()+
  labs(title = "GISAD Chilean Genomes",
       x = "Date",
       y = "# Covid Genomes",
       fill = "Variants") +
  theme(text = element_text(size = 12))
  #theme(text = element_text(size = 12), legend.position = "bottom")
p3
```

We compose all plots to generate figure 1

```{r cmposer}

library(patchwork)

patchwork=p1|(p2/p3)  
patchwork + plot_annotation(tag_levels = 'A')
pdf(file="Fig1.pdf",height=7,width=9)
patchwork + plot_annotation(tag_levels = 'A')
dev.off()

```


# Figure 2

Philogenetics, variants detected and coverage.

## Phylogenetic tree
```{r readdata,echo=FALSE}
#BiocManager::install("treeio")
#BiocManager::install("ggtree")

suppressMessages(library('treeio'))
suppressMessages(library('ggtree'))
suppressMessages(library(phytools))

all_results <- read.table(file='all_results.csv', sep=',', h=TRUE)

my_tree <- read.newick('../Phylogenetics/iqtree2.treefile')
rerooted_tree <- reroot(my_tree, node.number = 1)
#transforming tree to add variant information
x <- as_tibble(rerooted_tree)
#formatting sample name
x$sampleId <- paste0('sample_',(str_extract(x$label,"(?<=-)\\d+")))
#keeping name of reference
x <- x %>% 
  mutate(sampleId = ifelse(label == 'MN908947',label,sampleId))
#joining data
x_new <- treeio::full_join(x,all_results[c('sampleId', 'location', 'variant')], by="sampleId")
#transforming data into a tree object again!
x_new <- as.treedata(x_new)
#we rename the tips
a=x_new@phylo$tip.label
a=paste0('sample_',(str_extract(a,"(?<=-)\\d+")))
a[1]="MN908947"
x_new@phylo$tip.label=a
variant <- setNames(RColorBrewer::brewer.pal(name = "Dark2", n = length(unique(all_results$variant))), unique(all_results$variant))

ivar_results <- as.matrix(read.table('ivar_results.tsv', 
                           h=TRUE,sep="\t", check.names = FALSE, row.names = 1))
#deleting sample 71
ivar_results <- ivar_results[-71,]

pos_order=as.character(sort(as.numeric(colnames(ivar_results))))
#sort by coordinate
ivar_sor=ivar_results[,pos_order]
ivdt=as.data.frame(ivar_sor)
ivdt$name=rownames(ivdt)
ivdt_longer=ivdt%>%pivot_longer(!name, names_to = "pos", values_to = "snp") %>% mutate(pos=as.numeric(pos)) %>% filter(snp==1)
a=rowSums(ivar_results)
ivar_c=data.frame(name=names(a),value=a)

```

We plot the tree including variants type, iVAR snp distribution and number of SNPs


```{r plott}

library(ggtree)

# Customize further as needed
p4 <- ggtree(x_new, color="grey40", ladderize = TRUE) +
  #xlim_tree(0.00325) +
  geom_tiplab(aes(color=variant, label = sampleId), size=2.) +
  scale_color_manual('', values = variant) +
  scale_fill_manual('', values = variant) +
  guides(color = guide_legend(override.aes = list(size = 5, label = "\u25AA")))+
  theme_tree2(legend.position = "bottom")

# add SNP data by position in the tree
p5= p4 + geom_facet(panel = "SNPs coordinates", data = ivdt_longer, geom = geom_point, 
               mapping=aes(x = pos,group=sampleId, color=variant,fill=variant), shape = '|') +
             geom_facet(panel = "# SNPs", data = ivar_c, geom = geom_col, 
                aes(x = value, color = variant, fill=variant), orientation = 'y', width = .6)
p5

#facet_widths(p8, widths = c(2, 2, 1))
pdf(file="Fig2.pdf",height=7,width=9)
facet_widths(p5, widths = c(2, 2, 1))
dev.off()
```




# Figure 3

Genome coverage

## Samtools depth 

Checking depth of the samples and separate them by variant and location.

```{r, loadsamtools, echo=TRUE}
suppressMessages(library('reshape2'))
samtools_results <-read.table('samtools_depth.csv', h=TRUE,sep=",")
samtools_results <- melt(samtools_results, id="X")
colnames(samtools_results) <- c('pos', 'sampleId', 'depth')

samtools_results <- left_join(samtools_results,all_results[c('sampleId', 'location', 'variant')], by="sampleId")
samtools_results<-samtools_results[!grepl("71", samtools_results$sample),]

samtools_results <- samtools_results %>% group_by(variant)

samtools_results <- samtools_results %>% group_by(variant) %>%
  mutate(med = mean(depth, na.rm = TRUE))

samtools_depth_range <- samtools_results %>%
  group_by(sampleId, grp = cut(pos, breaks=pretty(pos, n = 60), dig.lab = 5),  variant, med) %>% 
  summarise(count = mean(depth, na.rm = TRUE))
```

### Samtools depth plot
```{r, depth, echo=FALSE}



p6 <- ggplot(samtools_depth_range , aes(x = grp, y=log(count),  
                                     colour = variant)) +
  geom_line(aes(group = sampleId), linewidth = 0.3, alpha = 0.8)+
  #geom_hline(yintercept = log(100),color="red2",
   #          linetype='dotted',linewidth=0.5) + 
  #annotate("text", x = 45, y = log(100), label = "100X Coverage", vjust = -0.5, size=2) +
  ylim(8,14)+
  scale_x_discrete(labels=as.character(seq(0,30000,500)))+
  facet_grid(rows = vars(variant),scales="fixed") +
  scale_color_manual('', values = variant) +
  geom_hline(aes(yintercept = log(med), group=variant), 
             colour = 'darkblue', size =0.6, linetype='dashed') +
  geom_text(aes(y = log(med), x = 45, label = paste(round(med),"X Coverage"), 
                group=variant),
                vjust = -0.6, size=4, color='gray40') + 
    theme_minimal() + 
  xlab("Position range (pb)") + ylab("Depth (log)") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 0.5, size = 5), legend.position = "bottom")

p6

pdf(file="Fig3.pdf",height=7,width=7)
p6
dev.off()
```



# Figure 4

Heterogeneity analysis


```{r bdata}
library(tidyverse)
dh=read.table("summary_het2.txt",h=F)
dh$V1=paste0('sample_',(str_extract(dh$V1,"(?<=-)\\d+")))
dh=left_join(dh,all_results,by=c("V1"="sampleId"))
dh=dh %>% mutate(age2=if_else(age < 30,"young",if_else(age >=30 & age<50,"adult","older"))) %>%  mutate(variant=if_else(variant=="BA.2* [Omicron (BA.2.X)]","Omicron",variant)) %>%remove_missing()

#count the number of position with ITH
dh$het=dh$V4+dh$V6

#run the lm
a=lm(dh$het~dh$age2+dh$mut_ivar+dh$variant)
summary(a)

#ggplot(dh, aes(y=het,x=age2,color=variant,group=age2))  + geom_boxplot() + geom_quasirandom()

```